---
title: "Curve Registration: Practical Guide and Pitfalls"
output:
  rmarkdown::html_vignette:
    code_folding: hide
vignette: >
  %\VignetteIndexEntry{Curve Registration: Practical Guide and Pitfalls}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
  collapse = TRUE,
  comment = "#>"
)
library(tf)
par(las = 1)
alpha_palette <- c(
  "#00000080", "#DF536B80", "#61D04F80", "#2297E680", "#28E2E580",
  "#CD0BBC80", "#F5C71080", "#9E9E9E80", "#00000080", "#DF536B80"
)
have_srvf <- rlang::is_installed("fdasrvf")
have_fda <- rlang::is_installed("fda")
```

This vignette is written in two passes:

1. First pages: practical guidance for choosing methods, avoiding common mistakes, and diagnosing results.
2. Later pages: more detail on method assumptions and a deeper case study.

## 1) Before You Register: Template, Warp, Domain

Registration aligns functions in time by reducing phase variation (timing differences) while trying to preserve amplitude variation (shape/height differences).

Use registration when:

- Curves have broadly similar shapes but shifted or stretched timing.
- Your scientific question depends on timing alignment.
- You can interpret both the **target template** and the resulting warps in domain terms.

Do not treat registration as a default preprocessing step when:

- Curves come from clearly different shape classes/clusters.
- Timing differences are not meaningful for your question.
- Heavy warping would create implausible subject-specific timelines.

Core model:

\[
x_i(t) \approx m\left(h_i^{-1}(t)\right)
\]

- \(m\): template shape you register against.
- \(h_i\): warping function mapping aligned time \(s\) to observed time \(t\).
- \(h_i^{-1}\): inverse warping used by `tf_unwarp()` to align observed curves.

### 1.1) What Is the Template?

Template choice is the most important modeling choice in registration: every estimated warp is relative to that template.

Default template behavior in `tf_register()`:

| Method | Default template behavior | How to override |
|:--|:--|:--|
| `srvf` | Karcher-type mean estimated by `fdasrvf` | pass `template = ...` |
| `fda` | arithmetic mean curve | pass `template = ...` |
| `affine` | arithmetic mean curve | pass `template = ...` |
| `landmark` | column-wise mean of landmarks | pass `template_landmarks = ...` |

Practical rules:

1. If you need interpretable timing relative to a known reference, supply an explicit template.
2. Keep template and data on the same domain/grid for `srvf` and `fda`.
3. Always visualize raw curves and the template together before trusting warps.

### 1.2) What a Warping Function Means

`tf_register()` returns warps \(h_i\). In practice, you usually inspect inverse warps \(h_i^{-1}\) via `tf_invert(warp)` because those are directly used for alignment.

- Curves close to the identity line mean little phase correction.
- Strong local slope changes imply strong local time dilation/compression.
- Large departures from identity without clear alignment gain indicate over-warping risk.

### 1.3) Domain-Preserving vs Non-domain-preserving Warps

This distinction is fundamental for interpretation and edge behavior:

- **Domain-preserving**: warps map domain endpoints to themselves. Typical for `srvf`, `fda`, and `landmark`.
- **Non-domain-preserving**: warps can shift/scale outside (or into a subset of) the observed domain. Often seen with `affine`.

Implications:

1. Non-domain-preserving warps can produce boundary `NA`s after `tf_unwarp()` because parts of the aligned timeline have no observed support.
2. Domain-preserving warps avoid these boundary losses but can still overfit timing.
3. Method choice is therefore not only about alignment quality; it is also about what time transformation is scientifically admissible.

Core workflow in `tf`:

```r
warp <- tf_register(x, method = "...")
x_reg <- tf_unwarp(x, warp)
```

## 2) Five-Minute Quick Start (Template-aware Baseline)

This minimal example uses affine shift registration because it has no extra package dependency and provides a fast baseline. We pass an explicit template to make the target shape transparent.

```{r quickstart, class.source = "fold-show"}
pinch_small <- pinch[1:10]
template_affine <- mean(pinch_small)

warp_affine <- tf_register(
  pinch_small,
  method = "affine",
  template = template_affine,
  type = "shift"
)
pinch_small_reg <- tf_unwarp(pinch_small, warp_affine)
```

```{r quickstart-plot}
layout(t(1:3))
plot(
  pinch_small,
  main = "Original Curves",
  xlab = "Observed Time [s]",
  ylab = "Force [N]",
  lwd = 1.5,
  col = alpha_palette
)
lines(template_affine, lwd = 2.5, lty = 3)
plot(
  tf_invert(warp_affine),
  points = FALSE,
  main = "Inverse Warping Functions",
  xlab = "Observed Time [s]",
  ylab = "Registered Time [s]",
  lwd = 1.5,
  col = alpha_palette
)
abline(0, 1, lty = 3)
plot(
  pinch_small_reg,
  main = "Registered Curves",
  xlab = "Registered Time [s]",
  ylab = "Force [N]",
  lwd = 1.5,
  col = alpha_palette
)
```

Quick success checks:

- The template is plausible for the population you are aligning.
- Major features (e.g., peaks) are more aligned after registration.
- Warps are monotone and not excessively bent for this problem.
- Boundary `NA`s are limited and scientifically acceptable.
- Resulting timing shifts are domain-plausible.

## 3) Choosing a Registration Method

> **Provisional guidance:** Method-selection advice in Sections 3/4/5 reflects current package behavior and practical experience. We are running a benchmark study comparing methods under controlled scenarios, so these recommendations may be revised after those results are finalized.

| Method | Use when | Avoid when | Template handling | Key arguments | Typical failure signs | First fallback |
|:--|:--|:--|:--|:--|:--|:--|
| `srvf` (default) | Smooth, non-linear timing differences with shared shape family | Very noisy data, sparse grids, highly heterogeneous shapes | Implicit Karcher-type mean unless `template` supplied | `method = "srvf"` | Unstable warps, over-warping, inconsistent reruns | Smooth more; try `affine` or `landmark` |
| `fda` | You want continuous-criterion alignment and can tune criterion | Complex phase-amplitude interactions or severe irregularity | Arithmetic mean unless `template` supplied | `method = "fda"`, `crit = 1/2` | Sensitive to criterion choice, weak alignment gains | Try `srvf` or `affine` |
| `affine` | Mostly shift/scale timing variability; need interpretable warps | Strong non-linear timing deformation | Arithmetic mean unless `template` supplied | `method = "affine"`, `type`, bounds | Residual misalignment of local features, boundary `NA`s under stronger shifts/scales | Upgrade to `srvf` |
| `landmark` | Reliable, repeated landmarks are available across curves | Ambiguous/noisy landmarks or mismatched counts | Uses `template_landmarks` (default: column means) | `method = "landmark"`, `landmarks`, `template_landmarks` | Forced/broken alignments from bad landmarks | Use `srvf` or `affine` |

Practical default:

1. Start with `affine` to get a baseline.
2. Move to `srvf` if you need non-linear timing flexibility.
3. Use `landmark` when domain landmarks are trusted and central.

## 4) Pitfalls and Likely Failure Modes (Read Before Scaling Up)

> **Provisional guidance:** Method-selection advice in Sections 3/4/5 reflects current package behavior and practical experience. We are running a benchmark study comparing methods under controlled scenarios, so these recommendations may be revised after those results are finalized.

### A) Multiple shape families (non-identifiability)

- **Detect:** Registered curves still form distinct shape clusters.
- **Do next:** Stratify first; register within clusters; compare cluster-specific results.

### B) Noise dominates phase signal

- **Detect:** Warps become jagged or highly variable across near-identical curves.
- **Do next:** Smooth first; reduce basis flexibility; compare before/after sensitivity.

### C) Sparse or irregular grids

- **Detect:** Registration quality depends strongly on interpolation grid choices.
- **Do next:** Re-evaluate to a stable common grid; run sensitivity checks on grid density.

### D) Boundary and overlap artifacts

- **Detect:** `NA` regions after unwarping, especially for aggressive affine shifts/scales.
- **Do next:** Inspect boundary behavior explicitly; use `keep_new_arg = TRUE` when needed; narrow affine bounds.

### E) Landmark quality issues

- **Detect:** Small landmark perturbations cause large warp changes.
- **Do next:** Validate landmarks visually; use robust landmark definitions; fall back to `srvf`/`affine` when ambiguous.

### F) Over-warping

- **Detect:** Large timing distortions with little gain in feature alignment.
- **Do next:** Prefer simpler method (`affine`), reduce flexibility, or avoid registration for that subset.

### G) Template mismatch

- **Detect:** Registered curves align to a shape that is not scientifically meaningful for your question.
- **Do next:** Choose an explicit template (or template landmarks), rerun, and compare conclusions.

## 5) Practical Diagnostics Workflow

> **Provisional guidance:** Method-selection advice in Sections 3/4/5 reflects current package behavior and practical experience. We are running a benchmark study comparing methods under controlled scenarios, so these recommendations may be revised after those results are finalized.

After any registration run:

1. Plot original curves together with the template you used.
2. Plot registered curves and compare key features to the template.
3. Plot inverse warps and check for plausible deformation.
4. Quantify one alignment target (e.g., spread of peak locations).
5. Check boundary support (`NA` rates), especially for affine warps.
6. Re-run with at least one alternative method.
7. Re-run after modest smoothing/grid change; check stability.

```{r diagnostics, class.source = "fold-show"}
x <- pinch[1:10]

warp_aff <- tf_register(x, method = "affine", type = "shift")
x_aff <- tf_unwarp(x, warp_aff)

peak_before <- as.numeric(tf_landmarks_extrema(x, "max"))
peak_after_aff <- as.numeric(tf_landmarks_extrema(x_aff, "max"))

data.frame(
  metric = c("sd_peak_before", "sd_peak_after_affine"),
  value = c(sd(peak_before), sd(peak_after_aff))
)

na_prop_aff <- mean(vapply(tf_evaluations(x_aff), \(v) mean(is.na(v)), numeric(1)))
na_prop_aff
```

```{r diagnostics-optional-srvf, eval = have_srvf, class.source = "fold-show"}
warp_srvf <- tf_register(x, method = "srvf")
x_srvf <- tf_unwarp(x, warp_srvf)
peak_after_srvf <- as.numeric(tf_landmarks_extrema(x_srvf, "max"))

data.frame(
  metric = c("sd_peak_after_affine", "sd_peak_after_srvf"),
  value = c(sd(peak_after_aff), sd(peak_after_srvf))
)
```

If different methods disagree strongly and diagnostics are unstable, treat conclusions as low confidence until you resolve data representation and model-choice sensitivity.

## 6) Method Notes and Theoretical Background

Registration models observed curves as time-deformed versions of a template:

\[
x_i(t) \approx m(h_i^{-1}(t))
\]

with monotone warping \(h_i\).

Key interpretation:

- \(h_i\) maps aligned/system time to observed time.
- \(h_i^{-1}\) maps observed time to aligned time.
- Slope \(>1\) or \(<1\) indicates local dilation/compression in time.

Method-specific notes:

- **SRVF**: strongest general-purpose option for smooth non-linear timing effects; sensitive to representation/noise.
- **FDA criteria**: useful alternatives but often require more tuning.
- **Affine**: constrained but interpretable and fast; excellent baseline.
- **Landmark**: highly effective when landmarks are reliable; fragile otherwise.

## 7) Worked Examples (Later Detail)

### 7.1 Pinch Data: Cross-method Comparison

```{r pinch-compare, class.source = "fold-show"}
pinch_small <- pinch[1:10]

warp_aff <- tf_register(pinch_small, method = "affine", type = "shift")
reg_aff <- tf_unwarp(pinch_small, warp_aff)

peak_locs <- tf_landmarks_extrema(pinch_small, "max")
warp_lm <- tf_register(pinch_small, method = "landmark", landmarks = peak_locs)
reg_lm <- tf_unwarp(pinch_small, warp_lm)
```

```{r pinch-compare-plot, fig.width = 8, fig.height = 8}
layout(t(matrix(1:4, 2, 2)))
plot(pinch_small, main = "Original", col = alpha_palette, lwd = 1.5)
plot(reg_aff, main = "Affine Registered", col = alpha_palette, lwd = 1.5)
plot(reg_lm, main = "Landmark Registered", col = alpha_palette, lwd = 1.5)
plot(tf_invert(warp_aff), points = FALSE, main = "Affine Inverse Warps", col = alpha_palette, lwd = 1.5)
abline(0, 1, lty = 3)
```

```{r pinch-compare-srvf-fda, eval = have_srvf || have_fda, class.source = "fold-show"}
if (have_srvf) {
  warp_srvf <- tf_register(pinch_small, method = "srvf")
  reg_srvf <- tf_unwarp(pinch_small, warp_srvf)
}
if (have_fda) {
  warp_fda <- tf_register(pinch_small, method = "fda", crit = 1)
  reg_fda <- tf_unwarp(pinch_small, warp_fda)
}
```

### 7.2 Growth Data: Deeper Real-data Workflow

```{r growth-prep-short, class.source = "fold-show"}
growth <- tf::growth
growth$growth <- tfb(
  growth$height,
  k = 20,
  bs = "ps",
  arg = seq(1, 18, l = 80),
  verbose = FALSE
) |>
  tf_derive()

if (have_srvf) {
  growth_warp <- tf_register(growth$growth, method = "srvf")
} else {
  growth_warp <- tf_register(growth$growth, method = "affine", type = "shift")
}
growth_reg <- tf_unwarp(growth$growth, growth_warp)
```

```{r growth-plot-short}
layout(t(1:3))
plot(
  growth$growth,
  xlab = "Chronological Age [years]",
  ylab = "Growth [cm/year]",
  lwd = 1.2
)
plot(
  tf_invert(growth_warp),
  points = FALSE,
  xlab = "Chronological Age",
  ylab = "Biological Age",
  lwd = 1.2
)
abline(0, 1, lty = 3)
plot(
  growth_reg,
  xlab = "Biological Age [years]",
  ylab = "Growth [cm/year]",
  lwd = 1.2
)
```

For larger studies, keep a diagnostics log per run (data preprocessing, method, key parameters, alignment metrics, and failure notes) so method choices are auditable and reproducible.

## 8) Default Playbook (Copy/Paste)

1. Start with `affine` as a baseline and inspect results.
2. Move to `srvf` if non-linear timing misalignment remains.
3. Use `landmark` only with high-confidence landmarks.
4. Run diagnostics from Section 5 every time.
5. Treat recommendations in Sections 3/4/5 as provisional pending benchmark updates.
