---
title: "curve registration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{registration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Functional data registration is a preprocessing technique that aligns functional
observations by removing phase variation (timing differences) while preserving amplitude
variation (shape differences). This process helps reveal the underlying patterns in
functional data by synchronizing features across observations.

The registration process involves:

1. Computing warping functions using `tf_register()` - these capture the timing differences between functions.
2. Applying the inverse warping using `tf_unwarp()` - this aligns the functions to a common time scale.

Let's explore this with real-world datasets.

## Growth Data

Growth curves from children often exhibit similar shapes but with different timing of
growth spurts. Registration can help identify common growth patterns by aligning these
developmental milestones.

```{r growth-prep}
# library(tf)
devtools::load_all(".")

growth_female <- subset(tf::growth, gender == "female", select = height)
growth_female$height_mean <- mean(growth_female$height)
growth_female$growth_mean <- tf_derive(growth_female$height_mean)
growth_female$growth <- tf_derive(growth_female$height)
growth_female$growth_warp <- tf_register(growth_female$growth)
growth_female$growth_reg <- tf_unwarp(
  growth_female$growth,
  growth_female$growth_warp
)
n <- nrow(growth_female)
```

### Original Growth Rate Trajectories

First, let's examine the unregistered growth rate curves. Notice how the timing of growth
spurts varies significantly across individuals:

```{r growth-original}
par(mfrow = c(1, 1))
plot(
  growth_female$growth,
  main = "Female Growth Rate Trajectories",
  xlab = "Age (years)",
  ylab = "Growth Rate (cm/year)",
  lwd = 1.5,
  col = seq_len(n),
  bty = "l",
  las = 1
)
```

### Warping Functions

The warping functions show how each individual's chronological age maps to a common "biological age". Deviations from the diagonal line indicate accelerated or delayed development:

```{r growth-warping}
plot(
  growth_female$growth_warp,
  xlab = "Clock Year",
  ylab = "Biological Year",
  lwd = 1.5,
  col = seq_len(n),
  bty = "l",
  las = 1,
)
abline(0, 1, lty = 2)
grid()
```

### Registered Growth Curves

After registration, the growth spurts are aligned, revealing the common underlying 
growth pattern:

```{r growth-aligned}
plot(
  growth_female$growth_reg,
  main = "Registered Female Growth Curves",
  xlab = "Age (years)",
  ylab = "Growth Rate (cm/year)",
  lwd = 1.5,
  col = seq_len(n),
  bty = "l",
  las = 1
)
grid()
```

### Individual Comparisons

Let's examine each case individually to see the effect of registration:

```{r growth-individuals}
oldpar <- par(mfrow = c(1, 2), pty = "s", ask = FALSE)
on.exit(par(oldpar))
for (i in seq_len(nrow(growth_female))) {
  plot(
    growth_female[i, "growth"],
    xlim = c(0, 18),
    main = paste("Growth Case", i),
    xlab = "Age (years)",
    ylab = "Growth Rate (cm/year)",
    lwd = 1.5,
    col = 4,
    bty = "l",
    las = 1
  )
  lines(growth_female$growth_mean, lty = 2, col = 2)
  lines(growth_female[i, "growth_reg"], col = 3)
  plot(
    growth_female[i, "growth_warp"],
    xlab = "Clock Year",
    ylab = "Biological Year",
    lwd = 1.5,
    bty = "l",
    las = 1,
  )
  abline(0, 1, lty = 2)
}
```

*FS: hier würde ich dann eher nur paar interessant Individuen auswählen, und ne legende für die farben brauchts auch...*

## Gait Data

```{r gait}
gait$knee_angle_warp <- tf_register(gait$knee_angle)
gait$hip_angle_warp <- tf_register(gait$hip_angle)
gait$knee_angle_reg <- tf_unwarp(gait$knee_angle, gait$knee_angle_warp)
gait$hip_angle_reg <- tf_unwarp(gait$hip_angle, gait$hip_angle_warp)

n <- nrow(gait)

par(mfrow = c(1, 1))
plot(
  gait$hip_angle,
  main = "Hip Angles During Gait Cycle",
  xlab = "Gait Cycle (%)",
  ylab = "Hip Angle (degrees)",
  lwd = 1.5,
  col = seq_len(n),
  bty = "l",
  las = 1
)
grid()

plot(
  gait$hip_angle_warp,
  main = "Hip Angle Warping Functions",
  xlab = "Original Gait Cycle (%)",
  ylab = "Registered Gait Cycle (%)",
  lwd = 1.5,
  col = seq_len(n),
  bty = "l",
  las = 1,
)
abline(0, 1, lty = 2)
grid()

plot(
  gait$hip_angle_reg,
  main = "Registered Hip Angles",
  xlab = "Gait Cycle (%)",
  ylab = "Hip Angle (degrees)",
  lwd = 1.5,
  col = seq_len(n),
  bty = "l",
  las = 1
)
grid()

oldpar <- par(mfrow = c(1, 2), pty = "s", ask = FALSE)
on.exit(par(oldpar))
for (i in seq_len(n)) {
  plot(
    gait[i, "hip_angle"],
    main = paste("Hip Angle Case", i),
    xlab = "Gait Cycle (%)",
    ylab = "Hip Angle (degrees)",
    lwd = 1.5,
    col = 4,
    bty = "l",
    las = 1
  )
  lines(gait$hip_mean, lty = 2, col = 2)
  lines(gait[i, "hip_angle_reg"], col = 3)

  plot(
    gait[i, "hip_angle_warp"],
    main = paste("Warping Case", i),
    xlab = "Original Gait Cycle (%)",
    ylab = "Registered Gait Cycle (%)",
    lwd = 1.5,
    bty = "l",
    las = 1,
  )
  abline(0, 1, lty = 2)
}

var_pre <- tf_fvar(gait$hip_angle)
var_post <- tf_fvar(gait$hip_angle_reg)
var_red <- sum(var_post) / sum(var_pre)
print(sprintf("Variance reduction hip angle: %.2f%%", 100 * (1 - var_red)))

var_pre <- tf_fvar(gait$knee_angle)
var_post <- tf_fvar(gait$knee_angle_reg)
var_red <- sum(var_post) / sum(var_pre)
print(sprintf("Variance reduction knee angle: %.2f%%", 100 * (1 - var_red)))
```

## Pinch Data

The pinch dataset contains biomechanical measurements of force exerted 
during a brief pinching motion by the thumb and forefinger. In this 
experiment, subjects maintained a constant background force of 2 newtons and 
then briefly squeezed to reach a target of 10 newtons before returning to 
the background level.

This dataset illustrates a common challenge in functional data analysis: the 
timing of events (in this case, when the pinch begins) varies arbitrarily 
across observations. Registration can align these force trajectories by 
synchronizing the timing of the pinching motion.

```{r pinch-prep}
pinch_warp <- tf_register(pinch)
pinch_reg <- tf_unwarp(pinch, pinch_warp)
```

### Original Pinch Force Trajectories

The unregistered data shows clear variation in when the pinching motion begins relative to the recording start time:

```{r}
plot(
  pinch,
  main = "Pinch Force Trajectories",
  xlab = "Time (s)",
  ylab = "Force (N)",
  lwd = 1.5,
  col = seq_along(pinch),
  bty = "l",
  las = 1
)
```

###  Pinch Warping Functions

The warping functions reveal the timing adjustments needed to align the pinching motions:

```{r pinch-warping}
plot(
  pinch_warp,
  xlab = "Original Time (s)",
  ylab = "Registered Time (s)",
  lwd = 1.5,
  col = seq_along(pinch),
  bty = "l",
  las = 1
)
```

### Registered Pinch Force Trajectories

After registration, the pinching motions are temporally aligned, revealing the common force pattern:

```{r pinch-registered}
plot(
  pinch_reg,
  main = "Registered Pinch Force Trajectories",
  xlab = "Time (s)",
  ylab = "Force (N)",
  lwd = 1.5,
  col = seq_along(pinch),
  bty = "l",
  las = 1
)
```