---
title: "curve registration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{registration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Functional data registration is a preprocessing technique that aligns functional
observations by removing phase variation (timing differences) while preserving amplitude
variation (shape differences). This process helps reveal the underlying patterns in
functional data by synchronizing features across observations.

The registration process involves:

1. Computing warping functions using `tf_register()` - these capture the timing differences between functions.
2. Applying the inverse warping using `tf_unwarp()` - this aligns the functions to a common time scale.

Let's explore this with real-world datasets.

## Growth Data

Growth curves from children often exhibit similar shapes but with different timing of
growth spurts. Registration can help identify common growth patterns by aligning these
developmental milestones.

```{r growth-prep}
library(tf)

growth_female <- subset(tf::growth, gender == "female", select = height)
growth_female$height_mean <- mean(growth_female$height)
growth_female$growth_mean <- tf_derive(growth_female$height_mean)
growth_female$growth <- tf_derive(growth_female$height)
growth_female$growth_warp <- tf_register(growth_female$growth)
growth_female$growth_reg <- tf_unwarp(
  growth_female$growth,
  growth_female$growth_warp
)
n <- nrow(growth_female)
```

### Original Growth Rate Trajectories

First, let's examine the unregistered growth rate curves. Notice how the timing of growth
spurts varies significantly across individuals:

```{r growth-original}
plot(
  growth_female$growth,
  main = "Female Growth Rate Trajectories",
  xlab = "Age (years)",
  ylab = "Growth Rate (cm/year)",
  lwd = 1.5,
  col = seq_len(n),
  bty = "l",
  las = 1
)
grid(col = "gray90", lty = "solid", lwd = 0.5)
```

### Warping Functions

The warping functions show how each individual's chronological age maps to a common "biological age". Deviations from the diagonal line indicate accelerated or delayed development:

```{r growth-warping}
plot(
  growth_female$growth_warp,
  xlab = "Clock Year",
  ylab = "Biological Year",
  lwd = 1.5,
  col = seq_len(n),
  bty = "l",
  las = 1,
)
abline(0, 1, lty = 2)
grid(col = "gray90", lty = "solid", lwd = 0.5)
```

### Registered Growth Curves

After registration, the growth spurts are aligned, revealing the common underlying 
growth pattern:

```{r growth-aligned}
plot(
  growth_female$growth_reg,
  main = "Registered Female Growth Curves",
  xlab = "Age (years)",
  ylab = "Growth Rate (cm/year)",
  lwd = 1.5,
  col = seq_len(n),
  bty = "l",
  las = 1
)
grid(col = "gray90", lty = "solid", lwd = 0.5)
```

### Individual Comparisons

We've chosen three cases that demonstrate distinct warping function behaviors and 
their effects on growth registration:

**Case 37 (Early Developer):**

The warping function curves below the diagonal identity line, indicating accelerated 
development. This child's growth spurt occurs earlier in chronological time - notice how 
the original orange curve shows peak growth around age 1-2, while the population mean 
peaks later. The warping function compresses the early years (steep rise from 0-2 years)
then gradually catches up to normal timing. Registration successfully aligns this 
early peak with the population pattern.

**Case 38 (Delayed Developer):** 

The warping function shows a dramatic S-curve pattern, staying well below the 
identity line until around age 8, then rapidly accelerating. This represents 
significantly delayed development - the original growth curve shows very little 
early growth followed by a late, intense growth spurt around age 10-12.
The warping function reveals how this child's "biological clock" runs much 
slower initially, requiring substantial time compression during the late growth phase.

**Case 32 (Variable Development):**

This case shows a more complex warping pattern with moderate deviations from the 
identity line. The warping function indicates slightly delayed early development 
followed by periods of acceleration and deceleration. The original growth curve 
shows multiple smaller peaks rather than one dominant spurt, and registration 
helps align these variable patterns with the smoother population average.

These examples illustrate how registration reveals individual differences in 
developmental timing while preserving the underlying growth patterns common 
to all children.

```{r growth-individuals}
green_col <- rgb(0.2, 0.4, 0.1, 0.7)
blue_col <- rgb(0.3, 0.5, 0.8, 0.7)
orange_col <- rgb(0.8, 0.4, 0.1, 0.7)

for (i in c(37, 38, 32)) {
  plot(
    growth_female[i, "growth"],
    xlim = c(0, 18),
    main = paste("Growth Case", i),
    xlab = "Age (years)",
    ylab = "Growth Rate (cm/year)",
    lwd = 1.5,
    col = orange_col,
    bty = "l",
    las = 1
  )
  lines(growth_female$growth_mean, lty = 2, col = green_col)
  lines(growth_female[i, "growth_reg"], col = blue_col)
  grid(col = "gray90", lty = "solid", lwd = 0.5)
  legend(
    "topright",
    legend = c("original", "mean", "registered"),
    col = c(orange_col, green_col, blue_col),
    pch = c(16, 17, 15),
    bty = "n",
    pt.cex = 1.4,
    cex = 1.0,
    text.col = "black",
    horiz = FALSE,
    inset = c(-0.1, 0)
  )

  plot(
    growth_female[i, "growth_warp"],
    main = paste("Warping Function Case", i),
    xlab = "Clock Year",
    ylab = "Biological Year",
    lwd = 1.5,
    bty = "l",
    las = 1,
  )
  abline(0, 1, lty = 2)
  grid(col = "gray90", lty = "solid", lwd = 0.5)
}
```

## Gait Data

```{r gait}
gait$knee_angle_warp <- tf_register(gait$knee_angle)
gait$hip_angle_warp <- tf_register(gait$hip_angle)
gait$knee_angle_reg <- tf_unwarp(gait$knee_angle, gait$knee_angle_warp)
gait$hip_angle_reg <- tf_unwarp(gait$hip_angle, gait$hip_angle_warp)

n <- nrow(gait)

par(mfrow = c(1, 1))
plot(
  gait$hip_angle,
  main = "Hip Angles During Gait Cycle",
  xlab = "Gait Cycle (%)",
  ylab = "Hip Angle (degrees)",
  lwd = 1.5,
  col = seq_len(n),
  bty = "l",
  las = 1
)
grid()

plot(
  gait$hip_angle_warp,
  main = "Hip Angle Warping Functions",
  xlab = "Original Gait Cycle (%)",
  ylab = "Registered Gait Cycle (%)",
  lwd = 1.5,
  col = seq_len(n),
  bty = "l",
  las = 1,
)
abline(0, 1, lty = 2)
grid()

plot(
  gait$hip_angle_reg,
  main = "Registered Hip Angles",
  xlab = "Gait Cycle (%)",
  ylab = "Hip Angle (degrees)",
  lwd = 1.5,
  col = seq_len(n),
  bty = "l",
  las = 1
)
grid()

oldpar <- par(mfrow = c(1, 2), pty = "s", ask = FALSE)
for (i in seq_len(n)) {
  plot(
    gait[i, "hip_angle"],
    main = paste("Hip Angle Case", i),
    xlab = "Gait Cycle (%)",
    ylab = "Hip Angle (degrees)",
    lwd = 1.5,
    col = 4,
    bty = "l",
    las = 1
  )
  lines(gait$hip_mean, lty = 2, col = 2)
  lines(gait[i, "hip_angle_reg"], col = 3)

  plot(
    gait[i, "hip_angle_warp"],
    main = paste("Warping Case", i),
    xlab = "Original Gait Cycle (%)",
    ylab = "Registered Gait Cycle (%)",
    lwd = 1.5,
    bty = "l",
    las = 1,
  )
  abline(0, 1, lty = 2)
}
par(oldpar)
```

## Pinch Data

The pinch dataset contains biomechanical measurements of force exerted
during a brief pinching motion by the thumb and forefinger. In this
experiment, subjects maintained a constant background force of 2 newtons and
then briefly squeezed to reach a target of 10 newtons before returning to
the background level.

This dataset illustrates a common challenge in functional data analysis: the
timing of events (in this case, when the pinch begins) varies arbitrarily
across observations. Registration can align these force trajectories by
synchronizing the timing of the pinching motion.

```{r pinch-prep}
pinch_warp <- tf_register(pinch)
pinch_reg <- tf_unwarp(pinch, pinch_warp)
```

### Original Pinch Force Trajectories

The unregistered data shows clear variation in when the pinching motion begins relative to the recording start time:

```{r}
plot(
  pinch,
  main = "Pinch Force Trajectories",
  xlab = "Time (s)",
  ylab = "Force (N)",
  lwd = 1.5,
  col = seq_along(pinch),
  bty = "l",
  las = 1
)
```

###  Pinch Warping Functions

The warping functions reveal the timing adjustments needed to align the pinching motions:

```{r pinch-warping}
plot(
  pinch_warp,
  xlab = "Original Time (s)",
  ylab = "Registered Time (s)",
  lwd = 1.5,
  col = seq_along(pinch),
  bty = "l",
  las = 1
)
```

### Registered Pinch Force Trajectories

After registration, the pinching motions are temporally aligned, revealing the common force pattern:

```{r pinch-registered}
plot(
  pinch_reg,
  main = "Registered Pinch Force Trajectories",
  xlab = "Time (s)",
  ylab = "Force (N)",
  lwd = 1.5,
  col = seq_along(pinch),
  bty = "l",
  las = 1
)
```

## Simulation Study

```{r simulation-study, eval = FALSE}
library(checkmate)
library(dplyr)
library(purrr)
library(tf)

#' Integrated Root Mean Squared Error
#'
#' @details
#' The Integrated Root Mean Squared Error is defined as \deqn{
#'   \sqrt{\int\left(\hat{f}_i(t)-f_i(t)\right)^2 d t}
#' }
tf_irmse <- function(truth, response) {
  assert_class(truth, "tfd")
  assert_class(response, "tfd")
  sqrt(tf_integrate((response - truth)^2))
}

tf_rgam <- function(n, arg = 51L, sigma = 0.1) {
  rlang::check_installed("fdasrvf")
  n <- assert_count(n, positive = TRUE, coerce = TRUE)
  arg <- assert_count(arg, positive = TRUE, coerce = TRUE)
  assert_number(sigma)
  x <- fdasrvf::rgam(N = arg, sigma = sigma, num = n)
  tfd(x, arg = seq(0, 1, length.out = arg))
}

# DGP 1: only phase variation
# DGP 2: rank-1 amplitude + phase
# DGP 3: complex phase (3‐basis mix)

# function recovery
# step 1. data generating process (20-50 replications)
# step 2. apply warping to simulated curves
# step 3. apply registration to warped curves
# step 4. compute IRMSE between original and registered curves

dgp_phase_only <- function(n = 20L, arg = 51L, sigma = 0.1) {
  x <- seq(-3, 3, length.out = 51)
  x <- sapply(seq_len(n), function(i) {
    amplitude <- runif(1, 0.8, 1.2)
    baseline <- runif(1, 0, 0.2)
    amplitude * sin(x) + baseline
  })
  x <- as.tfd(t(x))

  domain <- tf_domain(x)
  warp <- tf_rgam(n, arg = arg, sigma = sigma)
  warp <- domain[1] + (domain[2] - domain[1]) * warp
  warp <- tfd(tf_evaluations(warp), arg = tf_arg(x))
  unreg <- tf_warp(x, warp)
  list(reg = x, unreg = unreg, warp = warp)
}

dgp_rank_one <- function() {}

dgp_complex <- function() {}

run_recovery <- function(data, params) {
  pmap(params, function(...) {
    pv <- list(...)
    time <- system.time({
      warp_pred <- do.call(tf_register, c(list(.x = data$unreg), pv))
    })
    x_pred <- tf_unwarp(data$unreg, warp_pred)
    method <- pv$.method
    pv$.method <- NULL
    tibble(
      method = method,
      params = list(pv),
      mean = tf_irmse(mean(data$reg), mean(x_pred)),
      fn = mean(tf_irmse(data$reg, x_pred)),
      warp = mean(tf_irmse(data$warp, warp_pred)),
      time = unname(time["elapsed"])
    )
  })
}

run_benchmark <- function(params, ..., iter = 20L) {
  sim <- dgp_phase_only(...)
  bmr <- map(seq_len(iter), function(i) {
    run_recovery(sim, params) |>
      list_rbind() |>
      mutate(iter = i, .before = 1L)
  })
  list_rbind(bmr)
}

params <- expand.grid(
  penalty_method = c("roughness", "geodesic"),
  centroid_type = c("mean", "median"),
  stringsAsFactors = FALSE
) |>
  mutate(.method = "srvf", parallel = TRUE)

params <- data.frame(crit = 1:2, .method = "fda")

bmr_one <- run_benchmark(params, sigma = 0.1, iter = 3L)
bmr_two <- run_benchmark(params, sigma = 0.3, iter = 20L)
```