---
title: "Curve Registration"
output: 
  rmarkdown::html_vignette:
    code_folding: hide
vignette: >
  %\VignetteIndexEntry{Curve Registration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
  collapse = TRUE,
  comment = "#>"
)
library(tf)
par(las = 1)
# scales::alpha(1:10, .5)
alpha_palette <- c("#00000080", "#DF536B80", "#61D04F80", "#2297E680", "#28E2E580", 
"#CD0BBC80", "#F5C71080", "#9E9E9E80", "#00000080", "#DF536B80")

```

> **Note (new practical vignette):** For a more accessible, action-first guide with explicit method selection and diagnostic workflow, see `registration2.Rmd`. Method-selection advice is currently provisional and may be revised after the ongoing benchmark study comparing registration methods.

Functional data registration is a pre-processing technique that *aligns* functional observations by removing phase variation (timing differences) while preserving amplitude variation (shape differences). Misalignment of curves makes it difficult to distinguish between these two sources of variability. Registration separates phase and amplitude variation, enabling a clearer analysis of structural differences among curves.

The basic idea is that, in the simplest case, the observed curves $x_i(t)$ observed at times $t$ are elastically deformed versions of a common template function $m(s)$ with system time $s$: $x_i(t) = m(h_i(s)) = m(h_i^{-1}(t))$, with a strictly monotone increasing *warping function* $h_i(s)$ that maps the original "system" time argument $s$ of the template function to the observed "wall-clock" time argument $t$ of $x_i(t)$. Where the slope of $h(s)$ is below 1, observed time is sped up relative to the system time (i.e., the graph of the template function gets squished together in the horizontal direction), where the slope of $h(s)$ is above 1, observed time is slowed down relative to the system time (i.e., the graph of the template function is stretched out in the horizontal direction).  
Where the value of $h(s)$ is below the bisecting angle, the observed function is ahead of the template function (i.e., the observed function's features occur earlier than in the template), where $h(s)$ is above the bisecting angle, the observed function is behind the template function (i.e., the observed function's features occur later than in the template).  
Its inverse $h_i^{-1}(t)$ can be called *registration function*, *aligning function* or *unwarping  function* and the interpretation of its slope and position relative to the bisecting angle is reversed.   
For example:
```{r,  message = FALSE, fig.width = 8, fig.height = 8, fig.cap="Example warping and unwarping functions: Colored functions in the left column are warped versions of the template on the left. Middle column shows warping functions that deform the template into the observed functions on top and the inverse warpings that align the functions on the common template on the bottom.", out.width = "100%"}
set.seed(1221L)
# generate "template" function shape:
template <- tf_rgp(1, arg = 201L, nugget = 0)
# generate random warping functions (monotone increasing, [0, 1] -> [0, 1]):
# h_i(s) = \int_0^s exp(f_i(t)) dt / \int exp(f_i(s)) ds
warp <- {
   tmp <- tf_rgp(5)
   tmp <- exp(tmp - mean(tmp)) # center at identity warping
   tf_integrate(tmp, definite = FALSE) / tf_integrate(tmp)
}
x <- tf_warp(rep(template, 5), warp)

layout(cbind(c(0, 1, 1, 0), c(2, 2, 3, 3), c(0, 4, 4, 0)))
plot(template, 
     main = "Template / Registered Function", xlab = "System Time (aligned)")
plot(warp, col = 2:6, 
     main = "Warping Functions", xlab = "System Time", ylab = "Observed Time")
abline(c(0,1), col = "darkgrey", lty = 3)
plot(tf_invert(warp), col = 2:6, points = FALSE,
     main = "Inverse Warping Functions", xlab = "Observed Time", ylab = "System Time")
abline(c(0,1), col = "darkgrey", lty = 3)
plot(x, col = 2:6, 
     main = "Observed /\n Warped Functions", xlab = "Observed Time ('wall-clock')")
lines(template, lty = 3)
```

The registration process involves:

1.  Estimating "warping functions" using `tf_register()` - these capture the time scale deformations of the observed functions relative to a (typically iteratively estimated) common template.

2.  Applying the *inverse* warping using `tf_unwarp()` - this aligns the observed functions to the template by putting them on a common time scale.

The default registration method in `tf` uses the SRVF-framework implemented in the `fdasrvf` package, which is robust, mature, and has good theoretical properties. Alternatively, it also wraps the two "continuous criterion" registration methods implemented in the `fda` package, which are conceptually simpler but less theoretically appealing and less general (i.e., not recommended for data with complex phase- and amplitude variability). 

Some general caveats: 

- alignments can be fairly sensitive to the specifics of data representation (i.e., what kind of basis representation is being used, how many gridpoints the functions are being evaluated on,  how to interpolate between grid points, whether/how much the data was smoothed before registration). Even for a simple looking example like the "Berkeley Growth" data shown below, you can get very different alignment results from the same method, depending on the specific pre-processing.
- alignments can be fairly sensitive to the specific registration method being used and their hyperparameters. 
- registration problems are only well-defined and have a unique solution for functional datasets with simple amplitude variability^[c.f. [Chakraborty & Panaretos, 2020)](https://arxiv.org/pdf/1702.03556) for juicy details on this underappreciated fact]. If the curves being registered belong to different clusters with different mean shapes or just have very variable shapes more generally, the simple alignment methods implemented here are unlikely to yield sensible (or stable) results out of the box.  

Some general tips:

- smooth the data before aligning them -- especially the results of the SRVF method seem to be fairly sensitive to noise in the data.  
- evaluate curves on a dense enough grid to preserve all relevant small scale variation before aligning them. 
- for the SRVF method, equidistant grids seem to yield stable and sensible results more often than non-equidistant grids.

That said, let's explore how this works in `tf` with some real-world datasets:

## Pinch Data

The `pinch` data set contains bio-mechanical measurements of force exerted during a brief pinching motion by the thumb and forefinger. In this experiment, subjects maintained a constant background force of $2N$ and then briefly squeezed to reach a target of $10N$ before returning to the background level.

This data set illustrates a common challenge in functional data analysis: the timing of events (in this case, when the pinch begins and how long it takes to get to the peak and then relax again) varies across observations. Registration can align these force trajectories by synchronizing the timing of the pinching motion:
a global "system" time scale gets mapped to the individual time scales by the individual warping functions.

```{r pinch-prep, class.source = 'fold-show'}
library(tf)
pinch <- pinch[1:10]
pinch_warp <- tf_register(pinch) # get SRVF-warping functions (default method)
pinch_reg <- tf_unwarp(pinch, pinch_warp) # apply inverse warping to get registered functions
```

```{r pinch-viz}
layout(t(1:3))
plot(
  pinch,
  main = "Pinch Force Trajectories",
  xlab = "Observed Time [s]",
  ylab = "Force [N]",
  lwd = 1.5,
  col = alpha_palette,
)

plot(
  tf_invert(pinch_warp), #! show INVERSE warping, i.e. "alignment" functions
  points = FALSE,
  main = "Inverse Warping Functions",
  xlab = "Observed Time [s]",
  ylab = "Registered Time [s]",
  lwd = 1.5,
  col = alpha_palette
)
abline(0, 1, lty = 3, lwd = 2)

plot(
  pinch_reg,
  main = "Registered Pinch Force Trajectories",
  xlab = "Registered Time [s]",
  ylab = "Force [N]",
  lwd = 1.5,
  col = alpha_palette
)
``` 
The unregistered data (left panel) shows clear variation in when the pinching motion begins relative to the recording start time.
The unwarping functions `tf_invert(pinch_warp)` (middle panel) reveal the timing adjustments needed to align the pinching motions. We can see that some functions are pushed forward, some pushed backward in time at the beginning and 
the (un)warping functions then mostly have a slope $\approx 1$ (i.e., run parallel to the dotted bisector) for most of the time domain indicating no further time dilation or compression beyond the initial time shifts before re-converging towards the end.
After registration, the pinching motions are temporally aligned (right panel), revealing the common force pattern and its amplitude variation.

This is the kind of data with simple amplitude variability (i.e. the scaled bell curve shape all curves have in common) and straightforward phase variation (i.e., different on-sets and slightly different widths) for which registration/alignment tends to yield robust results.

Using either variant of the "continuous registration" methods yields fairly similar results here:
```{r pinch-prep2, class.source = 'fold-show', fig.width = 8, fig.height = 8}
pinch_warp_fda1 <- tf_register(pinch, method = "fda", crit = 1) # LS-criterion
pinch_warp_fda2 <- tf_register(pinch, method = "fda", crit = 2) # minEV-criterion
pinch_reg_fda1 <- tf_unwarp(pinch, pinch_warp_fda1)
pinch_reg_fda2 <- tf_unwarp(pinch, pinch_warp_fda2)

layout(t(matrix(1:4, 2, 2)))
plot(
  tf_invert(pinch_warp_fda1), 
  points = FALSE,
  main = "Inverse Warping Functions \n(fda; LS-criterion)",
  xlab = "Observed Time [s]",
  ylab = "Registered Time [s]",
  lwd = 1.5,
  col = alpha_palette
)
abline(0, 1, lty = 3, lwd = 2)
plot(
  pinch_reg_fda1,
  main = "Registered Pinch Force Trajectories",
  xlab = "Registered Time [s]",
  ylab = "Force [N]",
  lwd = 1.5,
  col = alpha_palette
)
plot(
  tf_invert(pinch_warp_fda2), 
  points = FALSE,
  main = "Inverse Warping Functions \n(fda; minEV-criterion)",
  xlab = "Observed Time [s]",
  ylab = "Registered Time [s]",
  lwd = 1.5,
  col = alpha_palette
)
abline(0, 1, lty = 3, lwd = 2)
plot(
  pinch_reg_fda2,
  main = "Registered Pinch Force Trajectories",
  xlab = "Registered Time [s]",
  ylab = "Force [N]",
  lwd = 1.5,
  col = alpha_palette
)
```

### Landmark Registration

When distinctive features (e.g., peaks, valleys, zero-crossings) are identifiable across curves, landmark registration provides a straightforward approach. For the pinch data, we can use the location of the peak force as a landmark to align the curves. `tf_landmarks_extrema()` finds the locations of extrema automatically:

```{r pinch-landmark, class.source = 'fold-show'}
# Find peak force locations for each curve
peak_locations <- tf_landmarks_extrema(pinch, "max")
peak_locations

# Register using these landmarks
pinch_warp_lm <- tf_register(pinch, method = "landmark", landmarks = peak_locations)
pinch_reg_lm <- tf_unwarp(pinch, pinch_warp_lm)
```

```{r pinch-landmark-viz}
layout(t(1:3))
plot(
  pinch,
  main = "Original Pinch Force",
  xlab = "Observed Time [s]",
  ylab = "Force [N]",
  lwd = 1.5,
  col = alpha_palette
)
# Mark the main peak (force maximum) for each curve
arg_vals <- as.numeric(tf_arg(pinch))
for (i in seq_along(pinch)) {
  evals <- tf_evaluations(pinch)[[i]]
  points(arg_vals[which.max(evals)], max(evals), pch = 19)
}

plot(
  tf_invert(pinch_warp_lm),
  points = FALSE,
  main = "Landmark Warping",
  xlab = "Observed Time [s]",
  ylab = "Registered Time [s]",
  lwd = 1.5,
  col = alpha_palette
)
abline(0, 1, lty = 3, lwd = 2)

plot(
  pinch_reg_lm,
  main = "Landmark-Registered",
  xlab = "Registered Time [s]",
  ylab = "Force [N]",
  lwd = 1.5,
  col = alpha_palette
)
```

Landmark registration produces piecewise linear warping functions that align the peaks. This is simpler than elastic registration and works well when the landmarks are reliable.

### Affine Registration

For data with primarily shift variability (or uniform scaling), affine registration provides a simpler and faster alternative. Affine warping functions have the form $h(t) = a \cdot t + b$, where $a$ controls scaling and $b$ controls shifting:

```{r pinch-affine, class.source = 'fold-show'}
# Shift-only registration (horizontal translation)
pinch_warp_shift <- tf_register(pinch, method = "affine", type = "shift")
pinch_reg_shift <- tf_unwarp(pinch, pinch_warp_shift)
```

```{r pinch-affine-viz}
layout(t(1:3))
plot(
  pinch,
  main = "Original Pinch Force",
  xlab = "Observed Time [s]",
  ylab = "Force [N]",
  lwd = 1.5,
  col = alpha_palette
)

plot(
  tf_invert(pinch_warp_shift),
  points = FALSE,
  main = "Affine (Shift) Warping",
  xlab = "Observed Time [s]",
  ylab = "Registered Time [s]",
  lwd = 1.5,
  col = alpha_palette
)
abline(0, 1, lty = 3, lwd = 2)

plot(
  pinch_reg_shift,
  main = "Shift-Registered",
  xlab = "Registered Time [s]",
  ylab = "Force [N]",
  lwd = 1.5,
  col = alpha_palette
)
```

Affine registration is appropriate when the phase variability is relatively simple. It's computationally faster than elastic registration and provides easily interpretable warping parameters.


## Case Study: Growth Data

Growth curves in children often display similar shapes but differ in the timing of growth spurts. Registration techniques can help reveal common growth patterns by aligning these developmental milestones. The `growth` dataset consists of longitudinal height measurements from 93 children in the Berkeley Growth Study, with data collected quarterly between ages 1 and 2, annually from 2 to 8 years, and biannually from 8 to 18 years. From these measurements, individual growth-rate curves are derived. We begin by registering the full set of height-velocity curves: we calculate each subject’s growth rate, compute the mean curve, estimate individual warping functions using `tf_register()`, and then align all curves to a common time scale with `tf_unwarp()`.

```{r growth-prep, class.source = 'fold-show'}
growth <- tf::growth
# for smoother, more accurate derivatives: 
# tfb on dense regular grid, using a basis w/o constraints on boundary deriv.
growth$growth <- 
  tfb(growth$height, 
      k = 20, bs = "ps", arg = seq(1, 18, l = 80),
      verbose = FALSE) |> 
  tf_derive() 
growth$growth_mean <- mean(growth$growth)
growth$growth_warp <- tf_register(growth$growth)
growth$growth_reg <- tf_unwarp(growth$growth, growth$growth_warp)
```

### Original Height Velocity Curves

First, let's examine the unregistered height velocity curves. Notice how the timing of growth spurts varies significantly across individuals:

```{r growth-original}
plot(
  growth$growth,
  xlab = "Age [years] ",
  ylab = "Growth [cm/year]",
  lwd = 1.5,
  bty = "l",
  las = 1
)
```

### Unwarping Functions

The unwarping functions show how each individual's chronological age maps to a common "biological age". Deviations from the diagonal line indicate accelerated or delayed development:

```{r growth-warping}
plot(
  tf_invert(growth$growth_warp),
  points = FALSE,
  xlab = "Chronological Age",
  ylab = "Biological Age",
  lwd = 1.5,
  bty = "l",
  las = 1
)
abline(c(0,1), lty = 3)
```

### Registered Growth Curves

After registration, the growth spurts are aligned, revealing the common underlying growth pattern:

```{r growth-aligned}
plot(
  growth$growth_reg,
  xlab = "Biological Age [years]",
  ylab = "Growth [cm/year]",
  lwd = 1.5,
  bty = "l",
  las = 1
)
```

### Growth by Biological Sex

Next, we stratify by sex (splitting the data into male and female groups) and repeat the same registration workflow within each subgroup—deriving sex‐specific mean growth rates, warping functions, and registered trajectories to directly compare male and female developmental timing and velocity^[We set `keep_new_arg = TRUE` in order to retain the registered `arg`-values instead of re-evaluating the aligned function on the old grid of the input function. This allows us to see how each observed datum is moved horizontally in the alignment.]

```{r growth-gender}
green_col <- rgb(0.2, 0.4, 0.1, 0.3)
orange_col <- rgb(0.8, 0.4, 0.1, 0.3)

growth_group <- growth |>
  split(~gender) |>
  lapply(function(growth) {
    growth$growth_mean <- mean(growth$growth)
    growth$growth_warp <- tf_register(growth$growth)
    # put unregistered & registered functions back on original grids for direct comparisons:
    growth$growth <- tfd(growth$growth, arg = tf_arg(growth$height))
    growth$growth_reg <- tf_unwarp(growth$growth, growth$growth_warp, keep_new_arg = TRUE) 
    growth
  })
growth_female <- growth_group$female
growth_male <- growth_group$male
```

```{r growth-gender-original}
plot(
  growth_male$growth,
  xlab = "Chronological Age [years]",
  ylab = "Height Velocity [cm/year]",
  lwd = 1.5,
  bty = "l",
  col = green_col,
  las = 1
)
lines(growth_female$growth, col = orange_col)
lines(growth_male$growth_mean, lty = 2)
lines(growth_female$growth_mean, lty = 2)
legend(
  "topright",
  legend = c("female", "male", "group means\n (unaligned)"),
  col = c(orange_col, green_col, "black"),
  pch = 16,
  bty = "n",
  pt.cex = 1.4,
  cex = 1.0,
  text.col = "black",
  horiz = FALSE
)
```

```{r growth-gender-warping}
plot(
  tf_invert(growth_male$growth_warp),
  points = FALSE,
  xlab = "Clock Year",
  ylab = "Biological Year",
  lwd = 1.5,
  bty = "l",
  col = green_col,
  las = 1
)
lines(tf_invert(growth_female$growth_warp), col = orange_col)
abline(0, 1, lty = 2)
```

```{r growth-gender-aligned}
plot(
  growth_male$growth_reg,
  arg = tf_arg(growth_male$growth_reg),
  points = FALSE,
  xlab = "Biological Age [years]",
  ylab = "Height Velocity [cm/year]",
  lwd = 1.5,
  bty = "l",
  col = green_col,
  las = 1
)
lines(growth_female$growth_reg, col = orange_col)

#rectify grids to compute mean
growth_male_mean_reg <- 
  tfd(growth_male$growth_reg, arg = tf_arg(growth_male$growth)) |> 
  mean() 
growth_female_mean_reg <- 
  tfd(growth_female$growth_reg, arg = tf_arg(growth_female$growth)) |> 
  mean() 

lines(growth_male_mean_reg, lty = 2)
lines(growth_female_mean_reg, lty = 2)

legend(
  "topright",
  legend = c("female", "male", "group means (registered)"),
  col = c(orange_col, green_col, "black"),
  pch = 16,
  bty = "n",
  pt.cex = 1.4,
  cex = 1.0,
  text.col = "black",
  horiz = FALSE
)
```

The results show that male children tend to have a more intense growth spurt later in puberty than female children.

###  Individual Comparisons

```{r growth-case-prep}
plot_patient_growth <- function(i) {
  plot(
    growth_female[i, "growth"],
    main = paste("Growth Case", i),
    xlab = "Age [years]",
    ylab = "Growth [cm/year]",
    lwd = 2,
    col = "red",
    bty = "l",
    las = 1,
  )
  arrows(x0 = tf_arg(growth_female[i, "growth"]), x1 = tf_arg(growth_female[i, "growth_reg"])[[1]],
         y0 = tf_evaluations(growth_female[i, "growth"])[[1]], y1 = tf_evaluations(growth_female[i, "growth"])[[1]],
        length = .1, angle = 20, col = rgb(0,0,0,.8)) |>     suppressWarnings()
  lines(growth_female[i, "growth_reg"], col = "blue", lwd = 2)

  legend(
    "topright",
    legend = c("original", "aligned"),
    col = c("red", "blue"),
    pch = 16,
    bty = "n",
    pt.cex = 1.4,
    cex = 1.0,
    text.col = "black",
    horiz = FALSE
  )
}

plot_patient_warp <- function(i) {
  plot(
    tf_invert(growth_female[i, "growth_warp"]),
    main = paste("Unwarping Function Case", i),
    xlab = "",
    ylab = "",
    lwd = 1.5,
    bty = "l",
    las = 1,
  )
  title(xlab = "Chronological Age (original)", col.lab = "red")
  title(ylab = "Biological Age (aligned)", col.lab = "blue")

  abline(0, 1, lty = 2)
}

# identify extreme and typical cases for viz:
# growth_female_warp_size <- tf_integrate(growth_female[, "growth_warp"])
# which(order(growth_female_warp_size) %in%  c(1, 54))
# most extreme warping:
# tf_derive(growth_female[, "growth_warp"])^2 |> tf_integrate() |> which.max()

```

We've chosen three female cases that demonstrate distinct warping function behaviors and their effects on height velocity registration:


**Case 5: Delayed Development**

```{r growth-case-5}
layout(t(1:2))
plot_patient_growth(5)
plot_patient_warp(5)
```
The left plot shows the original function $x_i(t)$ in red and its aligned version $x_i(h^-1_i(t))$ in blue, with points on the graph that are aligned to each other connected by horizontal arrows $t \to h^-1_i(t)$. The right plot shows the corresponding *unwarping* function $h^-1_i(t)$, the length of each arrow is the vertical distance of the unwarping function to the bisecting angle: the time shift at this point.

Development of this subject is aligned to the population mean pattern in the first 2-3 years
(unwarping function follows bisecting angle), and delayed over most of the rest of the observed period (unwarping function below bisecting angle, all arrows pointing left).
The graph of the original growth trajectory itself is compressed by the alignment early on (slope of unwarping function < 1 for age 2-7, arrows to the left lengthening), then dilated (steep unwarping function slope for age 7-10, arrows to the left shortening). 


**Case 38:  Complex Alignment**

```{r growth-case-35}
layout(t(1:2))
plot_patient_growth(38)
plot_patient_warp(38)
```

This case shows a more complex warping pattern with some strong deviation from the identity line. The warping function indicates some delayed early development (slower decrease in growth rates compared to population mean pattern), followed by a period of catch-up and then strong acceleration (age 3-8), followed by another period slower development.

<!-- TODO: include high level summary of simulation results here -->
