---
title: "Curve Registration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Curve Registration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Functional data registration is a preprocessing technique that aligns
functional observations by removing phase variation (timing differences) while
preserving amplitude variation (shape differences). Misalignment of curves
Registration separates phase and amplitude variation, enabling a clearer
makes it difficult to distinguish between these two sources of variability.
analysis of structural differences among curves.

The registration process involves:

1. Computing warping functions using `tf_register()` - these capture the timing
   differences between functions.
2. Applying the inverse warping using `tf_unwarp()` - this aligns the functions
   to a common time scale.

Let's explore this with real-world datasets.

## Pinch Data

The pinch dataset contains biomechanical measurements of force exerted during a
brief pinching motion by the thumb and forefinger. In this experiment, subjects
maintained a constant background force of 2 newtons and then briefly squeezed
to reach a target of 10 newtons before returning to the background level.

This dataset illustrates a common challenge in functional data analysis: the
timing of events (in this case, when the pinch begins) varies arbitrarily
across observations. Registration can align these force trajectories by
synchronizing the timing of the pinching motion.

```{r pinch-prep}
library(tf)

pinch_warp <- tf_register(pinch)
pinch_reg <- tf_unwarp(pinch, pinch_warp)
```

### Original Pinch Force Trajectories

The unregistered data shows clear variation in when the pinching motion begins
relative to the recording start time:

```{r pinch-original}
plot(
  pinch,
  main = "Pinch Force Trajectories",
  xlab = "Time (s)",
  ylab = "Force (N)",
  lwd = 1.5,
  col = seq_along(pinch),
  bty = "l",
  las = 1
)
```

###  Pinch Warping Functions

The warping functions reveal the timing adjustments needed to align the
pinching motions:

```{r pinch-warping}
plot(
  pinch_warp,
  xlab = "Original Time (s)",
  ylab = "Registered Time (s)",
  lwd = 1.5,
  col = seq_along(pinch),
  bty = "l",
  las = 1
)
abline(0, 1, lty = 2)
```

### Registered Pinch Force Trajectories

After registration, the pinching motions are temporally aligned, revealing the
common force pattern:

```{r pinch-aligned}
plot(
  pinch_reg,
  main = "Registered Pinch Force Trajectories",
  xlab = "Time (s)",
  ylab = "Force (N)",
  lwd = 1.5,
  col = seq_along(pinch),
  bty = "l",
  las = 1
)
```

## Growth Data

Growth curves in children often display similar shapes but differ in the timing
of growth spurts. Registration techniques help reveal common growth patterns by
aligning these developmental milestones. The growth dataset consists of
longitudinal height measurements from 93 children in the Berkeley Growth Study,
with data collected quarterly between ages 1 and 2, annually from 2 to 8 years,
and biannually from 8 to 18 years. From these measurements, individual
growth-rate curves are derived. We begin by registering the full set of
height-velocity curves: we calculate each subject’s growth rate, compute the
mean curve, estimate individual warping functions using `tf_register()`, and
then align all curves to a common time scale with `tf_unwarp()`.

```{r growth-prep}
growth$growth <- tf_derive(growth$height)
growth$growth_mean <- mean(growth$growth)
growth$growth_warp <- tf_register(growth$growth)
growth$growth_reg <- tf_unwarp(growth$growth, growth$growth_warp)
```

### Original Height Velocity Curves

First, let's examine the unregistered height velocity curves. Notice how the
timing of growth spurts varies significantly across individuals:

```{r growth-original}
plot(
  growth$growth,
  xlab = "Age (years)",
  ylab = "Height Velocity (cm/year)",
  lwd = 1.5,
  bty = "l",
  las = 1
)
```

### Warping Functions

The warping functions show how each individual's chronological age maps to a
common "biological age". Deviations from the diagonal line indicate accelerated
or delayed development:

```{r growth-warping}
plot(
  growth$growth_warp,
  xlab = "Clock Year",
  ylab = "Biological Year",
  lwd = 1.5,
  bty = "l",
  las = 1
)
```

### Registered Growth Curves

After registration, the growth spurts are aligned, revealing the common
underlying growth pattern:

```{r growth-aligned}
plot(
  growth$growth_reg,
  xlab = "Age (years)",
  ylab = "Height Velocity (cm/year)",
  lwd = 1.5,
  bty = "l",
  las = 1
)
```

Next, we stratify by gender (splitting the data into male and female groups)
and repeat the same registration workflow within each subgroup—deriving
gender‐specific mean growth rates, warping functions, and registered
trajectories to directly compare male and female developmental timing and
velocity

```{r growth-gender}
green_col <- rgb(0.2, 0.4, 0.1, 0.7)
orange_col <- rgb(0.8, 0.4, 0.1, 0.7)
black_col <- rgb(0.1, 0.1, 0.1, 0.7)

growth$growth <- tf_derive(growth$height)
growth_group <- growth |>
  split(~gender) |>
  lapply(function(growth) {
    growth$growth_mean <- mean(growth$growth)
    growth$growth_warp <- tf_register(growth$growth)
    growth$growth_reg <- tf_unwarp(growth$growth, growth$growth_warp)
    growth
  })
growth_female <- growth_group$female
growth_male <- growth_group$male
```

```{r growth-gender-original}
plot(
  growth_male$growth,
  xlab = "Age (years)",
  ylab = "Height Velocity (cm/year)",
  lwd = 1.5,
  bty = "l",
  col = green_col,
  las = 1
)
lines(growth_male$growth_mean, lty = 2, col = black_col)
lines(growth_female$growth, col = orange_col)
lines(growth_female$growth_mean, lty = 2, col = black_col)
legend(
  "topright",
  legend = c("female", "male", "mean"),
  col = c(orange_col, green_col, black_col),
  pch = 16,
  bty = "n",
  pt.cex = 1.4,
  cex = 1.0,
  text.col = "black",
  horiz = FALSE
)
```

```{r growth-gender-warping}
plot(
  growth_male$growth_warp,
  xlab = "Clock Year",
  ylab = "Biological Year",
  lwd = 1.5,
  bty = "l",
  col = green_col,
  las = 1
)
lines(growth_female$growth_warp, col = orange_col)
abline(0, 1, lty = 2)
```

```{r growth-gender-aligned}
plot(
  growth_male$growth_reg,
  xlab = "Age (years)",
  ylab = "Height Velocity (cm/year)",
  lwd = 1.5,
  bty = "l",
  col = green_col,
  las = 1
)
lines(growth_male$growth_mean, lty = 2, col = black_col)
lines(growth_female$growth_reg, col = orange_col)
lines(growth_female$growth_mean, lty = 2, col = black_col)
legend(
  "topright",
  legend = c("female", "male", "mean"),
  col = c(orange_col, green_col, black_col),
  pch = 16,
  bty = "n",
  pt.cex = 1.4,
  cex = 1.0,
  text.col = "black",
  horiz = FALSE
)
```

### Individual Comparisons

```{r growth-case-prep}
green_col <- rgb(0.2, 0.4, 0.1, 0.7)
blue_col <- rgb(0.3, 0.5, 0.8, 0.7)
orange_col <- rgb(0.8, 0.4, 0.1, 0.7)

plot_patient_growth <- function(i) {
  plot(
    growth_female[i, "growth"],
    main = paste("Growth Case", i),
    xlab = "Age (years)",
    ylab = "Height Velocity (cm/year)",
    lwd = 1.5,
    col = orange_col,
    bty = "l",
    las = 1
  )
  lines(growth_female[i, "growth_mean"], lty = 2, col = green_col)
  lines(growth_female[i, "growth_reg"], col = blue_col)
  legend(
    "topright",
    legend = c("original", "mean", "aligned"),
    col = c(orange_col, green_col, blue_col),
    pch = 16,
    bty = "n",
    pt.cex = 1.4,
    cex = 1.0,
    text.col = "black",
    horiz = FALSE
  )
}

plot_patient_warp <- function(i) {
  plot(
    growth_female[i, "growth_warp"],
    main = paste("Warping Function Case", i),
    xlab = "Clock Year",
    ylab = "Biological Year",
    lwd = 1.5,
    bty = "l",
    las = 1,
  )
  abline(0, 1, lty = 2)
}
```

We've chosen three cases that demonstrate distinct warping function behaviors
and their effects on height velocity registration:

**Case 37 (Early Developer):**

```{r growth-case-37}
plot_patient_growth(37)
plot_patient_warp(37)
```

The warping function curves above the diagonal identity line, indicating
accelerated development. This child's growth spurt occurs earlier in
chronological time - notice how the original orange curve shows peak growth
around age 1-2, while the population mean peaks later. The warping function
compresses the early years (steep rise from 0-2 years) then gradually catches
up to normal timing. Registration successfully aligns this early peak with the
population pattern.

**Case 38 (Delayed Developer):**

```{r growth-case-38}
plot_patient_growth(38)
plot_patient_warp(38)
```

The warping function shows a dramatic S-curve pattern, staying well below the
identity line until around age 8, then rapidly accelerating. This represents
significantly delayed development - the original growth curve shows very little
early growth followed by a late, intense growth spurt around age 10-12. The
warping function reveals how this child's "biological clock" runs much slower
initially, requiring substantial time compression during the late growth phase.

**Case 32 (Variable Development):**

```{r growth-case-32}
plot_patient_growth(32)
plot_patient_warp(32)
```

This case shows a more complex warping pattern with moderate deviations from
the identity line. The warping function indicates slightly delayed early
development followed by periods of acceleration and deceleration. The original
growth curve shows multiple smaller peaks rather than one dominant spurt, and
registration helps align these variable patterns with the smoother population
average.

These individual cases demonstrate the importance of registration in
understanding growth patterns. Without it, case 37's early growth spurt and
case 38's delayed spurt would appear to be completely different patterns of
growth. However, registration reveals that these children follow the same
underlying growth trajectory, just at different times. Properly accounting for
individual timing differences through registration enables us to better
identify children who deviate from the normal growth pattern in terms of shape
rather than just timing.
