---
title: "Vector-valued functional data: gait cycles and hurricane tracks"
output:
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{Vector-valued functional data: gait cycles and hurricane tracks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.2,
  dpi  = 96,
  out.width = "100%"
)
set.seed(1)
```

A `tf` vector represents a sample of functions `f: R -> R`. Many real
measurement processes, though, produce several coupled signals that share a
common argument: the `(hip, knee)` joint-angle pair sampled across a gait
cycle, the `(longitude, latitude)` position of a hurricane sampled in time,
the `(x, y, z)` body coordinates of a moving animal. The natural object is a
**vector-valued function** `f: R -> R^d`, and `tf` represents a sample of
those with the `tf_mv` family: `tfd_mv` for raw evaluations and `tfb_mv` for
basis representations. Internally a `tf_mv` is a bundle of `d` ordinary
`tf` vectors, so the multivariate methods shown below reuse the existing
univariate numerical kernels by mapping over components.

This article uses two real datasets to put the API through its paces:

* **`tf::gait`** -- regularly sampled hip and knee angles for 39 boys across
  one gait cycle. *Question: how variable is "normal" gait, where in the
  cycle does that variation concentrate, and what are the dominant modes of
  variation?*
* **`dplyr::storms`** -- irregularly sampled position and intensity of every
  Atlantic tropical storm or hurricane from `r min(dplyr::storms$year)` to
  `r max(dplyr::storms$year)`. *Question: do stronger storms travel further
  and move faster than weak ones?*

```{r packages, message = FALSE}
library(tf)
library(dplyr)
```

# 1. A quick tour of vector-valued functional data

Before the case studies, this part introduces the objects: what a vector-valued
function *is*, the two classes that represent samples of them, how to construct
them, and the handful of geometric operations the case studies lean on.

## What is a vector-valued function?

A `tf` vector stores a sample of scalar functions $f: [a, b] \to \mathbb{R}$. A
**vector-valued** function bundles $d$ of them on a *shared* argument,

$$ f(t) = \big(f_1(t), \dots, f_d(t)\big), \qquad t \in [a, b], $$

so that as $t$ runs over the domain the value $f(t)$ traces a *path* through
$\mathbb{R}^d$. `tf` represents a sample of such paths with the `tf_mv` family.
Internally a `tf_mv` *is* a bundle of `d` ordinary `tf` vectors sharing one
argument, so every univariate verb -- evaluation, arithmetic, derivatives,
integration, smoothing, basis fitting -- extends component-wise for free.

Two kinds of variation run through everything below: **amplitude** (how large
the values get) and **phase** (*where in* $t$ the features happen). Much of the
work in functional data analysis is telling the two apart, and Section 2 is
largely about exactly that.

## The `tf_mv` classes: `tfd_mv` and `tfb_mv`

Like their univariate cousins `tfd`/`tfb`, vector-valued functions come in two
flavours:

* `tfd_mv` -- *raw evaluations*: each component is a `tfd`, i.e. values on a grid
  plus an interpolation rule.
* `tfb_mv` -- *basis representation*: each component is a `tfb` (spline or FPC),
  i.e. basis coefficients.

Components are named, and a few accessors reach into the bundle: `tf_ncomp()`
(number of components), `tf_components()` (the list of univariate `tf` vectors),
`tf_component(x, k)` (one component, by name or index), and `names()`.

## Constructing `tf_mv` objects

`tfd_mv()` accepts three interchangeable input layouts. We use the built-in
`gait` data throughout the gait case study -- hip and knee angles, in degrees,
for 39 boys, each sampled at 20 points across one gait cycle. The most direct
construction is a **named list** of univariate `tf` vectors:

```{r tour-list}
data(gait)
g <- tfd_mv(list(hip = gait$hip_angle, knee = gait$knee_angle))
g
```

The print-out shows `d = 2` components on a shared grid, with a sparkline pair
per subject. It behaves like a vector of 39 bivariate curves; component
accessors and bracket extraction expose the two levels separately -- curves on
rows, argument values on columns, components on the third array dimension:

```{r gait-anatomy}
tf_ncomp(g)
names(tf_components(g))
dim(g[1:3, c(0.1, 0.5, 0.9)])
class(g[, , component = "hip"])
```

The *same* object can be built from a **3-d array**
`[curve, arg, component]` -- the layout `as.matrix()` returns --

```{r tour-array}
arr <- as.matrix(g)            # [curve, arg, component]
dim(arr)
g_arr <- tfd_mv(arr, arg = tf_arg(g))
identical(as.matrix(g_arr), arr)
```

or from a **wide data frame** with one column per component (the long
`(id, arg, component, value)` schema is what `as.data.frame(unnest = TRUE)`
emits; `long = FALSE` gives the wide schema the constructor consumes):

```{r tour-df}
df <- as.data.frame(g, unnest = TRUE, long = FALSE)   # id, arg, hip, knee
head(df, 2)
g_df <- tfd_mv(df, id = "id", arg = "arg", value = c("hip", "knee"))
```

Each route also takes an optional `domain` and a per-component `evaluator`
(interpolation rule, default `tf_approx_linear`). A `tfd_mv` becomes a basis
object with `tfb_mv()`, choosing a `"spline"` or `"fpc"` basis; basis arguments
apply globally or per component when passed as a named list:

```{r tour-basis}
g_spline <- tfb_mv(g, basis = "spline", k = list(hip = 5, knee = 12),
                   verbose = FALSE)
g_spline
```

## Geometry on the bundle: speed, arc length, reparametrization

Because a `tf_mv` traces a path, the usual curve-geometry quantities are
available, and they return *univariate* summaries of the multivariate object:

* `tf_speed(f)` $= \lVert f'(t)\rVert$ -- pointwise speed, a `tfd`,
* `tf_arclength(f)` -- total path length, one number per curve,
* `tf_reparam_arclength(f)` -- the same path traversed at *unit speed*.

```{r tour-geom, fig.height = 3.6}
head(tf_arclength(g))
plot(tf_speed(g), alpha = 0.3, main = expression("pointwise speed " * "||" * f*"'"*(t) * "||"))
```

## Aligning vector-valued curves: three notions

When curves differ mainly in *timing*, we usually want to factor that out --
but "factor out" can mean several different things. There is a ladder of
increasingly aggressive operations, distinguished by *what each treats as a
nuisance*:

| operation | removes | keeps | `tf` entry point |
|---|---|---|---|
| arc-length reparametrization | the clock (parametrization only) | shape **and** size | `tf_reparam_arclength()` |
| shared time-warp alignment | phase, onto a common template | amplitude, size, orientation | `tf_register(., method = "srvf_mv")` |
| full elastic shape registration | phase **+** rotation **+** scale | shape only | `tf_register_shape()` |

The first re-labels time without picking a template; the second warps every
curve onto a common template but leaves the values untouched; the third
*additionally* rotates and rescales each curve, landing in a **shape space**
where only the geometric form remains. Section 2 walks this ladder on the gait
data -- and shows where the top rung is, and is not, the right tool.

# 2. How variable is a "normal" gait cycle?

We built the two-component object `g` above. The two natural views of these 39
curves are the **time-series**
(`type = "facet"`, one panel per component) and the **trajectory in phase
space** (`type = "trajectory"`, the default for two-component objects).

```{r gait-facet, fig.height = 3.6}
plot(g, type = "facet", alpha = 0.4)
```

```{r gait-traj, fig.height = 4.2}
plot(g, alpha = 0.4)   # type = "trajectory" by default for d = 2
```

The facet view shows that both joints flex twice per cycle; the trajectory
view collapses time and reveals the characteristic "butterfly" loop traced
out as the leg cycles through stance and swing. Both views suggest the
between-subject spread is far from uniform along the cycle.

## Pointwise mean and standard deviation

`mean()` and `sd()` are vctrs group generics and dispatch component-wise:
they each return a length-1 `tfd_mv`. Plot the raw curves with `plot.tf`,
then overlay the mean +/- 2 sd envelope with `lines.tf` -- all arithmetic
on the components is component-wise on `tf_mv`, so `mu + 2 * s` is itself
a `tfd_mv` and `tf_component(...)` gives the per-axis envelope:

```{r gait-mean-sd, fig.height = 3.8}
mu <- mean(g)
s  <- sd(g)

op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
for (k in seq_len(tf_ncomp(g))) {
  nm <- names(tf_components(g))[k]
  plot(tf_component(g, k), alpha = 0.25,
       ylab = nm, main = nm)
  lines(tf_component(mu, k),              col = "red", lwd = 2)
  lines(tf_component(mu + 2 * s, k),      col = "red", lwd = 1.2, lty = 2)
  lines(tf_component(mu - 2 * s, k),      col = "red", lwd = 1.2, lty = 2)
}
par(op)
```

The pointwise sd peaks roughly where the angle itself is changing fastest --
around the stance/swing transitions -- so most of the between-subject
variability lives in *timing* of those transitions, not in the angles
attained at rest.

## Most- and least-varied subjects

`tf_arclength()` measures the path length traced out in `(hip, knee)`-space.
A short orbit = a subject whose `(hip, knee)` excursions are small or who
moves the two joints in lock-step; a long orbit = a subject with
large-amplitude, less synchronised excursions.

```{r gait-arc, fig.height = 4.2}
arc <- tf_arclength(g)
extreme <- c(which.min(arc), which.max(arc))
extreme

plot(g, alpha = 0.15)
lines(g[extreme[1]], col = "steelblue", lwd = 2.2)
lines(g[extreme[2]], col = "firebrick", lwd = 2.2)
legend("bottomright", bty = "n",
       legend = c(sprintf("min arc length (%.0f deg)", arc[extreme[1]]),
                  sprintf("max arc length (%.0f deg)", arc[extreme[2]])),
       col = c("steelblue", "firebrick"), lwd = 2)
```

## Phase alignment with `tf_register`

The pointwise sd above conflates two very different kinds of
between-subject variability: *amplitude* (how high does the knee swing?)
and *phase* (at what fraction of the cycle does heel-strike happen?).
`tf_register()` on a `tf_mv` estimates **one shared time-warp per curve**
and applies it to all components jointly. The warp is derived from a
single univariate "reference signal" -- which component (or which derived
quantity) to use is the central modelling choice.

Three reasonable choices, each emphasising different events in the cycle:

```{r gait-register-fit, results = "hide"}
r_hip   <- tf_register(g, method = "cc", ref_component = "hip")
r_knee  <- tf_register(g, method = "cc", ref_component = "knee")
# any univariate function of x can serve as the reference signal -- here the
# tangential *speed* in (hip, knee)-space, combining both components
r_speed <- tf_register(g, method = "cc", ref_component = tf_speed)
```

### Aligned trajectories

`plot(..., type = "facet")` shows the aligned curves for each choice of
reference signal. The reference signal that was *used* to register a
sample shrinks most strongly, but the partner component can shrink or
expand depending on how phase-coupled the two are:

```{r gait-register-helpers, class.source = "fold-hide"}
plot_aligned <- function(reg, title) {
  plot(tf_aligned(reg), type = "facet", alpha = 0.3)
  mtext(title, side = 3, line = -1.3, outer = TRUE, font = 2)
}
```

```{r gait-register-traj, fig.width = 8.2, fig.height = 7.2}
op <- par(mfrow = c(3, 1), oma = c(0, 0, 0, 0))
plot_aligned(r_hip,   "ref = hip angle")
plot_aligned(r_knee,  "ref = knee angle")
plot_aligned(r_speed, "ref = speed = ||(hip', knee')||")
par(op)
```

The "ref = knee" panel visibly tightens the knee component. Hip variation
changes much less, which is useful to see: a single shared warp can improve
the component that carries the sharpest event without necessarily removing
all variability in the smoother partner signal.

### Estimated warping functions

The warps themselves are univariate `tfd`s (one per subject), mapping the
raw cycle phase to the aligned phase. `plot.tf` handles them directly:

```{r gait-register-warps, fig.width = 8.2, fig.height = 3.6}
warps_hip   <- tf_invert(tf_inv_warps(r_hip))
warps_knee  <- tf_invert(tf_inv_warps(r_knee))
warps_speed <- tf_invert(tf_inv_warps(r_speed))

op <- par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
plot(warps_hip,   alpha = 0.5, main = "warp from hip",   ylab = "aligned phase")
abline(0, 1, lty = 3)
plot(warps_knee,  alpha = 0.5, main = "warp from knee",  ylab = "aligned phase")
abline(0, 1, lty = 3)
plot(warps_speed, alpha = 0.5, main = "warp from speed", ylab = "aligned phase")
abline(0, 1, lty = 3)
par(op)
```

The dotted diagonal is the identity warp (no time distortion). The
spread around the diagonal tells you the magnitude and pattern of the
phase variability captured by each reference signal: warps from `knee`
and `speed` bunch tightly near identity in the first half of the cycle
and fan out around the swing phase, where heel-strike timing varies
between subjects. The `hip`-driven warps are noticeably broader
elsewhere -- the hip signal is too smooth to lock the alignment to a
single sharp event.

### Quantifying the alignment

Compare the maximum pointwise sd before and after each registration:

```{r gait-register-table-helper, class.source = "fold-hide"}
peak_sd <- function(f, k) {
  arg <- tf_arg(f)
  max(unlist(tf_evaluate(tf_component(sd(f), k), arg)))
}
```

```{r gait-register-table}
out <- data.frame(
  reference = c("raw (no reg)", "hip", "knee", "speed"),
  sd_hip = c(peak_sd(g, "hip"),
             peak_sd(tf_aligned(r_hip),   "hip"),
             peak_sd(tf_aligned(r_knee),  "hip"),
             peak_sd(tf_aligned(r_speed), "hip")),
  sd_knee = c(peak_sd(g, "knee"),
              peak_sd(tf_aligned(r_hip),   "knee"),
              peak_sd(tf_aligned(r_knee),  "knee"),
              peak_sd(tf_aligned(r_speed), "knee"))
)
out
```

`knee` is the most informative reference signal for the sharp heel-strike
event: it substantially reduces knee variation, while hip variation is mostly
unchanged. We carry the knee-aligned sample forward because it best targets
the event where phase variation is most visible.

## Modes of variation via FPC

With the most visible phase variation reduced, fit a per-component FPC basis.
The leading PCs now mostly describe within-component amplitude modes:

```{r gait-fpc-helper, class.source = "fold-hide"}
fpc_var <- function(comp) attr(comp, "score_variance")
```

```{r gait-fpc, fig.height = 3.8}
g_aligned <- tf_aligned(r_knee)
g_b <- tfb_mv(g_aligned, basis = "fpc", verbose = FALSE)

v_hip  <- fpc_var(tf_component(g_b, "hip"));  v_hip  <- v_hip  / sum(v_hip)
v_knee <- fpc_var(tf_component(g_b, "knee")); v_knee <- v_knee / sum(v_knee)

op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
barplot(head(v_hip,  6), names.arg = 1:6, main = "hip: variance share",
        ylab = "proportion", col = "grey70")
barplot(head(v_knee, 6), names.arg = 1:6, main = "knee: variance share",
        ylab = "proportion", col = "grey70")
par(op)

# residual: how well does the low-rank FPC basis approximate the curves?
g_round <- vctrs::vec_cast(g_b, g_aligned)
resid   <- g_aligned - g_round
rmse_per_subject <- sqrt(unlist(lapply(tf_evaluations(resid), function(m) {
  mean(as.matrix(m[, -1L, drop = FALSE])^2)
})))
summary(rmse_per_subject)
```

The first 2-3 FPCs explain most of the remaining within-component variance,
and the RMSE between the registered and FPC-reconstructed `(hip, knee)`
curves is well below a degree for most subjects.

## Three ways to factor out "nuisance" variation

The reference-signal registration above is *one* way to strip out phase
variability. It is worth seeing it next to its neighbours on the ladder from
Section 1, each treating a *different* feature of the curves as a nuisance. We
walk all three rungs on the same gait sample.

### Rung 1 -- arc-length reparametrization

`tf_reparam_arclength()` traverses each curve at *approximately constant speed
in its value space*. The trajectory in the `(hip, knee)` plane is unchanged --
it is the same set of points -- but the time axis is redistributed so that equal
time intervals cover equal arc length. No template and no other curve enter:
this is a property of each curve on its own. `tf_speed()` returns a plain `tfd`,
so `plot.tf` and `lines.tf` work directly:

```{r gait-reparam, fig.height = 4.0}
g_unit  <- tf_reparam_arclength(g)
sp_raw  <- tf_speed(g)
sp_unit <- tf_speed(g_unit)

op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
plot(sp_raw,  alpha = 0.4, ylim = c(0, max(c(
       unlist(tf_evaluations(sp_raw)),
       unlist(tf_evaluations(sp_unit))), na.rm = TRUE)),
     main = "raw parameterization", ylab = "speed (deg / cycle)")
lines(mean(sp_raw),  col = "firebrick", lwd = 2)
plot(sp_unit, alpha = 0.4, ylim = c(0, max(c(
       unlist(tf_evaluations(sp_raw)),
       unlist(tf_evaluations(sp_unit))), na.rm = TRUE)),
     main = "arc-length parameterization", ylab = "speed (deg / cycle)")
lines(mean(sp_unit), col = "firebrick", lwd = 2)
par(op)
```

The raw plot shows large per-subject speed swings -- joints flex quickly
through swing and stand almost still during stance. The reparameterized
plot is approximately flat: each curve is traversed at the same speed
throughout the cycle, equal to its per-subject arc length divided by the
cycle duration. This removes *parametrization* differences but keeps every
curve's shape and size intact.

### Rung 2 -- shared time-warp alignment (`srvf_mv`)

`method = "srvf_mv"` estimates one time-warp per curve onto a common template,
using *all* components jointly -- in contrast to `method = "cc"` above, which
derived the warp from a single reference signal. There is no `ref_component`:
the warp is the one that best aligns the joint `(hip, knee)` trajectories in the
elastic (square-root-velocity) sense. The values are only re-timed, never
rotated or rescaled.

```{r gait-srvfmv, message = FALSE, warning = FALSE, fig.height = 3.6}
reg_mv <- tf_register(g, method = "srvf_mv", max_iter = 2)
reg_mv

plot(tf_aligned(reg_mv), type = "facet", alpha = 0.3)
```

The estimated warps are univariate `tfd`s, recovered from the inverse warps with
`tf_invert()`. Every curve is aligned to the shared Karcher-mean template in
`tf_template(reg_mv)`:

```{r gait-srvfmv-warps, fig.height = 4.0}
warps_mv <- tf_invert(tf_inv_warps(reg_mv))
plot(warps_mv, alpha = 0.5, main = "srvf_mv warps", ylab = "aligned phase")
abline(0, 1, lty = 3)
```

Unlike rung 1 this uses the whole sample (curves are aligned *to each other*
through the template); like rung 1 it leaves amplitude untouched.

### Rung 3 -- full elastic shape registration, and why *not* on gait

`tf_register_shape()` goes one step further: on top of the time-warp it finds,
per curve, a **rotation** and a **scale** that best match the template. It
returns a `tf_shape_registration` with two extra accessors, `tf_rotations()`
and `tf_scales()`:

```{r gait-shape, message = FALSE, warning = FALSE}
reg_shape <- tf_register_shape(g, max_iter = 2)
reg_shape
round(tf_rotations(reg_shape)[, , 1], 3)    # rotation for subject 1
summary(as.numeric(tf_scales(reg_shape)))   # scales, relative to the template
```

The estimated rotations sit essentially at the identity and the scales bunch
tightly around 1 -- the method finds almost nothing to do. That is the correct
answer, and the cautionary point: **gait's two axes are not interchangeable.**
Hip angle and knee angle are distinct physical quantities in fixed units.
Rotating the `(hip, knee)` plane would mix them into meaningless linear
combinations, and rescaling would discard amplitude -- which here *is* the
signal of interest. Shape registration is simply the wrong tool for a bundle of
fixed-unit channels.

> **When is rung 3 the right tool?** When the components are *interchangeable
> spatial coordinates* and orientation and overall size are genuine nuisances:
> handwriting and gesture trajectories, animal movement paths, outlines of
> objects. It is *not* appropriate for fixed-unit channel bundles such as
> `(hip, knee)` angles, `(wind, pressure)`, or geographic `(long, lat)`.

### A planar shape where rung 3 *is* the right tool

To see shape registration do real work, here is a small synthetic example: a
single base curve $(t,\, t^2)$, copied three times, with each copy rotated,
rescaled and shifted in the plane. Now orientation and size *are* nuisances and
only the common shape should survive.

```{r shape-demo-build}
t      <- seq(0, 1, length.out = 51)
base   <- rbind(t, t^2)
scales <- c(1, 0.7, 1.3)
angles <- c(0, 0.4, -0.25)
offset <- rbind(c(0.2, -0.1), c(0.4, -0.2), c(0.6, -0.3))
beta   <- array(NA_real_, dim = c(3, length(t), 2),
                dimnames = list(c("a", "b", "c"), NULL, c("x", "y")))
for (i in 1:3) {
  rot   <- matrix(c(cos(angles[i]),  sin(angles[i]),
                    -sin(angles[i]), cos(angles[i])), nrow = 2)
  curve <- scales[i] * (rot %*% base) + matrix(offset[i, ], 2, length(t))
  beta[i, , 1] <- curve[1, ]
  beta[i, , 2] <- curve[2, ]
}
shapes <- tfd_mv(beta, arg = t)
```

```{r shape-demo-register, message = FALSE, warning = FALSE, fig.width = 8.2, fig.height = 4.2}
reg_p <- tf_register_shape(shapes, max_iter = 2)

op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
plot(shapes,            asp = 1, col = 1:3, lwd = 2,
     main = "three placements of one shape")
plot(tf_aligned(reg_p), asp = 1, col = 1:3, lwd = 2,
     main = "after shape registration")
par(op)
```

The three curves were drawn at different positions, orientations and sizes;
after registration they collapse onto essentially one curve. The recovered scale
factors track the injected sizes -- reported *relative to the template*, so a
curve drawn larger comes back with a factor below 1 (shrunk toward the template)
and a smaller one above 1:

```{r shape-demo-scales}
data.frame(curve           = c("a", "b", "c"),
           injected_scale  = scales,
           recovered_scale = round(as.numeric(tf_scales(reg_p)), 3))
```

# 3. Atlantic storms as 4-dimensional curves

`dplyr::storms` records four time-varying quantities for every Atlantic
tropical storm or hurricane from `r min(dplyr::storms$year)` to
`r max(dplyr::storms$year)`, sampled every 6 hours along the storm's life:
position (longitude, latitude) and intensity (sustained wind speed, central
pressure). The natural object is a single
**four-component** vector-valued curve per storm, `f_i: [0, T_i] -> R^4`
-- spatial trajectory and intensity life-cycle bundled together.

Two practical points:

* `tracks_km` -- the *spatial* `(x_km, y_km)` view in physical units
  (projecting each storm's `long` / `lat` into a per-storm local-km
  frame at its own mean latitude) -- is what we use for arc length and
  forward speed, because deg/h speeds wrongly upweight east-west motion
  at high latitudes (a longitude degree is ~111 km at the equator but
  only ~78 km at 45 N).
* Most life-cycle comparisons between short and long storms only make
  sense on a **normalised time axis** `phase = t / T_i in [0, 1]`. We
  build a normalised 4-d object on top of the real-time one and use the
  appropriate version per question.

```{r storms-build, class.source = "fold-hide"}
KM_PER_DEG <- 111.32

storms_clean <- storms |>
  mutate(
    storm_id = paste(name, year),
    ts       = as.POSIXct(ISOdate(year, month, day, hour), tz = "UTC")
  ) |>
  group_by(storm_id) |>
  distinct(ts, .keep_all = TRUE) |>      # dedupe duplicate-timestamp rows
  mutate(
    t_hours = as.numeric(ts - min(ts), units = "hours"),
    phase   = if (max(t_hours) > 0) t_hours / max(t_hours) else 0,
    ref_lat = mean(lat),
    x_km    = (long - mean(long)) * KM_PER_DEG * cos(ref_lat * pi / 180),
    y_km    = (lat  - ref_lat)    * KM_PER_DEG
  ) |>
  filter(n() >= 16, !is.na(wind), !is.na(pressure)) |>  # >= 4 days
  ungroup()

# spatial (km, real time): the compact data-frame constructor.
tracks_km <- storms_clean |>
  transmute(storm_id, t_hours, x = x_km, y = y_km) |>
  tfd_mv(id = "storm_id", arg = "t_hours", value = c("x", "y"))

# Equivalent spelling if the component tfd vectors are already available.
tracks_km_from_components <- tfd_mv(list(
  x = tfd(storms_clean, id = "storm_id", arg = "t_hours", value = "x_km"),
  y = tfd(storms_clean, id = "storm_id", arg = "t_hours", value = "y_km")
))

# full 4-d, on normalised life-cycle phase
tracks4 <- tfd_mv(list(
  long = tfd(storms_clean, id = "storm_id", arg = "phase", value = "long"),
  lat  = tfd(storms_clean, id = "storm_id", arg = "phase", value = "lat"),
  wind = tfd(storms_clean, id = "storm_id", arg = "phase", value = "wind"),
  pres = tfd(storms_clean, id = "storm_id", arg = "phase", value = "pressure")
))
tracks4

# storm-level metadata
peak <- storms |>
  group_by(name, year) |>
  summarise(peak_cat = suppressWarnings(max(category, na.rm = TRUE)),
            .groups = "drop") |>
  mutate(peak_cat = if_else(is.finite(peak_cat), peak_cat, 0),
         peak_cat = as.integer(peak_cat),
         storm_id = paste(name, year))

df <- tibble::tibble(
  storm_id  = names(tracks4),
  track     = tracks4,
  track_km  = tracks_km
) |>
  left_join(peak, by = "storm_id") |>
  mutate(strength = factor(
    pmin(peak_cat, 4),
    levels = 0:4,
    labels = c("TS/TD", "Cat 1", "Cat 2", "Cat 3", "Cat 4+")
  ))
```

Before analysing the storms, inspect the irregular object directly. Counts
vary by storm because each lifetime has a different number of 6-hourly
observations, while the long data-frame representation keeps the components
aligned by `(storm, phase)`.

```{r storms-irregular}
head(tf_count(tracks4))
as.data.frame(tracks4[1:2], unnest = TRUE) |> head()
```

## Movement workflow: regularize, then describe

Joo et al. (2019) describe movement analysis as a workflow built around
tracking records `(x, y, t)`: clean the fixes, regularize or reconstruct
the path when needed, visualize the track, then extract descriptors such
as speed, heading and turning angles. The storm data follow the same
pattern. The build step above already performed basic preprocessing:
duplicate timestamps were removed, very short tracks were dropped, and
longitude/latitude were projected into local kilometre coordinates.

For diagnostic plots it is useful to regularize a few storms onto a
common lifecycle grid. Differentiating the regularized `(x, y)` curves
gives velocity components; their norm is forward speed, and `atan2(v_y,
v_x)` gives heading. Turning angle is the wrapped difference between
successive headings.

```{r storms-movement-workflow, fig.width = 8.2, fig.height = 6.6}
movement_ids <- c("Andrew 1992", "Katrina 2005", "Sandy 2012")
movement_ids <- movement_ids[movement_ids %in% df$storm_id]
movement_rows <- match(movement_ids, df$storm_id)
movement_grid <- seq(0, 1, length.out = 81)
movement_duration <- vapply(
  tf_arg(df$track_km[movement_rows]),
  \(t) max(t) - min(t),
  numeric(1)
)

movement_eval <- lapply(seq_along(movement_rows), function(k) {
  tf_evaluate(
    df$track_km[movement_rows[k]],
    arg = movement_grid * movement_duration[k]
  )[[1]]
})
movement_x <- do.call(rbind, lapply(movement_eval, `[[`, "x"))
movement_y <- do.call(rbind, lapply(movement_eval, `[[`, "y"))

movement_track <- tfd_mv(list(
  x = tfd(movement_x, arg = movement_grid, domain = c(0, 1)),
  y = tfd(movement_y, arg = movement_grid, domain = c(0, 1))
), domain = c(0, 1))
names(movement_track) <- movement_ids

movement_velocity <- tf_derive(movement_track)
movement_speed <- tf_norm(movement_velocity) / movement_duration

vx <- as.matrix(tf_component(movement_velocity, "x"),
                arg = movement_grid, interpolate = TRUE)
vy <- as.matrix(tf_component(movement_velocity, "y"),
                arg = movement_grid, interpolate = TRUE)
heading_mat <- atan2(vy, vx) * 180 / pi
heading <- tfd(heading_mat, arg = movement_grid, domain = c(0, 1))
names(heading) <- movement_ids

wrap_degrees <- function(x) ((x + 180) %% 360) - 180
turning_mat <- t(apply(heading_mat, 1, function(x) {
  c(NA_real_, wrap_degrees(diff(x)))
}))
turning <- tfd(turning_mat, arg = movement_grid, domain = c(0, 1))
names(turning) <- movement_ids

movement_cols <- c("#1b9e77", "#d95f02", "#7570b3")[seq_along(movement_ids)]
xy <- as.matrix(movement_track, interpolate = TRUE)

op <- par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
plot(range(xy[, , "x"], na.rm = TRUE), range(xy[, , "y"], na.rm = TRUE),
     type = "n", asp = 1, xlab = "x (km)", ylab = "y (km)",
     main = "regularized tracks")
lines(movement_track, col = movement_cols, lwd = 2)
legend("topleft", bty = "n", lwd = 2, col = movement_cols,
       legend = movement_ids, cex = 0.85)
plot(movement_speed, col = movement_cols, lwd = 2, alpha = 1,
     xlab = "lifecycle phase", ylab = "km / h", main = "forward speed")
plot(heading, col = movement_cols, lwd = 2, alpha = 1,
     xlab = "lifecycle phase", ylab = "degrees", main = "heading")
plot(turning, col = movement_cols, lwd = 2, alpha = 1,
     xlab = "lifecycle phase", ylab = "degrees", main = "turning angle")
par(op)
```

The result is a movement-data diagnostic rather than just a map: storms
with visually similar tracks can differ sharply in when they accelerate,
when they turn, and how abruptly their heading changes during recurvature.

## A single 4-d curve: Hurricane Katrina (2005)

For a single storm the 4-component `tf_mv` is naturally displayed
in facet mode (`type = "facet"`) -- one panel per component, all on the
same normalised cycle phase axis:

```{r storms-katrina, fig.width = 8, fig.height = 5.4}
katrina <- df |> filter(storm_id == "Katrina 2005") |> pull(track)
plot(katrina, type = "facet", lwd = 2)
```

Wind and pressure are anti-correlated -- pressure dips below 920 mbar as
wind peaks above 150 knots near phase ~0.55 -- and the `(long, lat)`
panels show Katrina sliding northwest across the Gulf and turning north
into Louisiana. Bundling intensity and trajectory as a single `tfd_mv`
lets every operation downstream (subsetting, summaries, arithmetic,
plotting) treat them as one object.

## Track map, faceted by peak intensity

The geographic view uses the `(long, lat)` components only. Faceting by
peak Saffir-Simpson category separates the populations cleanly, and a
coastline backdrop (from `maps::map`, if installed) anchors the geography:

```{r storms-plot-helpers, class.source = "fold-hide"}
have_maps <- requireNamespace("maps", quietly = TRUE)
pal <- c("grey50", "#fed976", "#feb24c", "#fd8d3c", "#e31a1c")

draw_coast <- function(xlim, ylim) {
  if (!have_maps) return(invisible())
  m <- maps::map("world", plot = FALSE,
                 xlim = xlim, ylim = ylim, fill = FALSE)
  lines(m$x, m$y, col = "grey55", lwd = 0.6)
}
```

```{r storms-map, fig.width = 8.2, fig.height = 6.8}
xlim <- range(unlist(tf_evaluations(tf_component(df$track, "long"))), na.rm = TRUE)
ylim <- range(unlist(tf_evaluations(tf_component(df$track, "lat"))),  na.rm = TRUE)

op <- par(mfrow = c(2, 3), mar = c(3.4, 3.4, 2, 1), mgp = c(2, 0.7, 0))
for (k in seq_along(levels(df$strength))) {
  lev   <- levels(df$strength)[k]
  trks  <- df$track[df$strength == lev]
  spatial <- tfd_mv(list(
    long = tf_component(trks, "long"),
    lat  = tf_component(trks, "lat")
  ))
  plot(range(xlim), range(ylim), type = "n",
       xlab = "long", ylab = "lat",
       main = sprintf("%s (n = %d)", lev, length(trks)))
  draw_coast(xlim, ylim)
  lines(spatial, col = pal[k], alpha = 0.4, lwd = 1)
}
plot.new()
legend("center", bty = "n", lwd = 2, col = pal, legend = levels(df$strength),
       title = "peak intensity", cex = 1.05)
par(op)
```

Tropical storms / depressions stay in the southern, western basin; cat-3
and cat-4+ tracks reach further north and east -- they live longer, get
caught by the westerlies, and recurve out to sea.

## Scalar features extracted from the 4-d object

Every component of a `tf_mv` is a real `tf` vector and supports the full
univariate API, so per-storm scalar features fall out as one-liners:

```{r storms-scalars}
df <- df |> mutate(
  path_km    = tf_arclength(track_km),
  duration   = vapply(tf_arg(track_km), \(t) max(t) - min(t), numeric(1)),
  mean_speed = path_km / duration,            # km/h, lifetime average
  peak_wind  = vapply(tf_evaluations(tf_component(track, "wind")), max, numeric(1)),
  min_pres   = vapply(tf_evaluations(tf_component(track, "pres")), min, numeric(1))
)

df |>
  group_by(strength) |>
  summarise(
    n = dplyr::n(),
    median_path_km     = round(median(path_km)),
    median_speed_kmh   = round(median(mean_speed), 1),
    median_peak_wind   = median(peak_wind),
    median_min_pres    = median(min_pres),
    .groups = "drop"
  )
```

Median path length grows strongly from tropical-storm to cat-4+ storms,
whereas lifetime-average forward speed increases more modestly. The
canonical wind / pressure gap between categories is visible.

## Intensity and forward-speed life-cycles per category

Plotting all forward-speed and intensity time courses at once is hopeless
(hundreds of irregular noodles). What we actually want is the
*mean curve per intensity stratum* on the normalised lifecycle phase
axis. Evaluate each component on a common phase grid, average within each
stratum, and re-wrap the result as a length-`G` univariate `tfd` so we
can use the standard `plot.tf` / `lines.tf` machinery:

```{r storms-lifecycle-prep, class.source = "fold-hide"}
phase_grid <- seq(0, 1, length.out = 41)

# forward speed on normalised time: re-arg each storm's km-speed by
# t / T_i, so all storms share phase domain [0, 1].
speed_km  <- tf_speed(df$track_km)
durations <- df$duration
speed_long <- do.call(rbind, lapply(seq_along(speed_km), function(i) {
  data.frame(
    id    = names(speed_km)[i],
    phase = tf_arg(speed_km[i])[[1]] / durations[i],
    value = tf_evaluations(speed_km[i])[[1]]
  )
}))
speed_phase <- tfd(speed_long, id = "id", arg = "phase", value = "value",
                   domain = c(0, 1))

# per-stratum mean curve on the regular phase grid, packaged as length-G tfd
stratum_mean_tfd <- function(comp, grp, grid = phase_grid) {
  mat <- as.matrix(comp, arg = grid, interpolate = TRUE)
  means <- vapply(levels(grp), function(g) {
    rows <- which(grp == g)
    if (length(rows) < 2L) return(rep(NA_real_, length(grid)))
    colMeans(mat[rows, , drop = FALSE], na.rm = TRUE)
  }, numeric(length(grid)))
  out <- tfd(t(means), arg = grid, domain = c(0, 1))
  names(out) <- levels(grp)
  out
}

wind_avg  <- stratum_mean_tfd(tf_component(df$track, "wind"), df$strength)
pres_avg  <- stratum_mean_tfd(tf_component(df$track, "pres"), df$strength)
speed_avg <- stratum_mean_tfd(speed_phase,                     df$strength)
```

```{r storms-lifecycle, fig.width = 8.2, fig.height = 6.8}
op <- par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
plot(wind_avg, col = pal, lwd = 2, alpha = 1,
     xlab = "lifecycle phase", ylab = "wind (knots)",
     main = "mean sustained wind")
legend("topright", bty = "n", lwd = 2, col = pal, legend = levels(df$strength),
       cex = 0.9, title = "peak intensity")
plot(pres_avg,  col = pal, lwd = 2, alpha = 1,
     xlab = "lifecycle phase", ylab = "pressure (mbar)",
     main = "mean central pressure")
plot(speed_avg, col = pal, lwd = 2, alpha = 1,
     xlab = "lifecycle phase", ylab = "forward speed (km/h)",
     main = "mean forward speed")
plot.new()
text(0.5, 0.7, "wind and pressure peak\nnear lifecycle phase 0.5;",
     adj = 0.5, cex = 1.05)
text(0.5, 0.35, "forward speed peaks\nlate, during recurvature",
     adj = 0.5, cex = 1.05)
par(op)
```

Three clean stories on a single `tf_mv` object: intensity peaks
mid-life-cycle (around `phase ~ 0.5`) and the gap between categories is
roughly uniform across the cycle; central pressure mirrors wind; forward
speed peaks much later (around `phase ~ 0.75`) and the strongest storms
accelerate the most -- the canonical recurvature signature.

## Smoothing on normalised time

Fitting a basis representation on raw real-time data has a tractable
domain-extrapolation problem -- the longest storm runs 600+ hours, but
most tracks end after 100-200 hours, and a shared basis over [0, 600]
extrapolates wildly. On the **normalised phase** version this just
disappears, because every storm spans `[0, 1]`:

```{r storms-tfb, fig.width = 8.2, fig.height = 4.8, warning = FALSE}
top6 <- df |> arrange(desc(path_km)) |> slice(1:6) |> pull(storm_id)

# fit a per-component spline basis on (long, lat, wind, pres) over [0, 1]
tb <- tfb_mv(df$track[df$storm_id %in% top6], k = 12, verbose = FALSE)

# pull just the (long, lat) components out of the 4-d objects so plot.tf_mv
# defaults to the trajectory (long, lat) view
raw_xy <- df$track[df$storm_id %in% top6, , c("long", "lat")]
sm_xy  <- tb[, , c("long", "lat")]

op <- par(mfrow = c(1, 2), mar = c(3.6, 3.6, 2, 1), mgp = c(2, 0.7, 0))
plot(range(xlim), range(ylim), type = "n",
     xlab = "long", ylab = "lat", main = "raw observations")
draw_coast(xlim, ylim)
lines(raw_xy, col = 1:6, lwd = 1.6)
plot(range(xlim), range(ylim), type = "n",
     xlab = "long", ylab = "lat", main = "spline-smoothed")
draw_coast(xlim, ylim)
lines(sm_xy, col = 1:6, lwd = 1.6)
par(op)
```

The smoothed trajectories pick up the gross recurvature shape without the
6-hourly sampling jitter and -- thanks to the normalised time axis --
without any extrapolation pathology. The same `tb` object also carries
clean smoothed wind / pressure curves for each storm; for example:

```{r storms-tfb-intensity, fig.width = 8.2, fig.height = 3.4}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
plot(tf_component(tb, "wind"), col = 1:6, lwd = 2,
     xlab = "lifecycle phase", ylab = "wind (knots)",
     main = "smoothed wind")
plot(tf_component(tb, "pres"), col = 1:6, lwd = 2,
     xlab = "lifecycle phase", ylab = "pressure (mbar)",
     main = "smoothed pressure")
par(op)
```

# 4. Recap

The two case studies above exercised exactly the same surface:

* construction from per-component `tf` vectors (`tfd_mv(list(...))`), from a 3-d
  `[curve, arg, component]` array, or from a long/wide data frame;
* `tf_components()`, `$component`, `tf_ncomp()` for component access;
* facet vs. trajectory plotting (`plot(..., type = ...)`);
* vctrs-native arithmetic / `mean()` / `sd()` returning a length-1 `tf_mv`;
* `tf_arclength()` and `tf_speed()` as geometric primitives on the bundle;
* `tfb_mv()` for smoothing (`basis = "spline"`) or PC decomposition
  (`basis = "fpc"`);
* the alignment ladder: `tf_reparam_arclength()` for constant-speed (shape-only)
  reparametrization, `tf_register()` for shared-warp alignment -- from a single
  reference component (`method = "cc"`) or jointly from all components
  (`method = "srvf_mv"`) -- and `tf_register_shape()` for full elastic shape
  registration, whose `tf_rotations()` / `tf_scales()` accessors expose the
  estimated rotations and (template-relative) scale factors;
* full dplyr / tibble integration via the underlying vctrs proxy.

The common pattern is simple: keep coupled signals bundled as one object, use
component accessors when a scalar or univariate summary is needed, and let the
multivariate methods reuse the existing univariate kernels component-wise.
