---
title: "Vector-valued functional data: gait cycles and hurricane tracks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vector-valued functional data: gait cycles and hurricane tracks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.2,
  dpi  = 96,
  out.width = "100%"
)
set.seed(1)
```

A `tf` vector represents a sample of functions `f: R -> R`. Many real
measurement processes, though, produce several coupled signals that share a
common argument: the `(hip, knee)` joint-angle pair sampled across a gait
cycle, the `(longitude, latitude)` position of a hurricane sampled in time,
the `(x, y, z)` body coordinates of a moving animal. The natural object is a
**vector-valued function** `f: R -> R^d`, and `tf` represents a sample of
those with the `tf_mv` family: `tfd_mv` for raw evaluations and `tfb_mv` for
basis representations. Internally a `tf_mv` is just a bundle of `d` ordinary
`tf` vectors, so every univariate verb -- evaluation, arithmetic, smoothing,
basis fitting, derivatives, integration -- extends component-wise for free.

This article uses two real datasets to put the API through its paces:

* **`tf::gait`** -- regularly sampled hip and knee angles for 39 boys across
  one gait cycle. *Question: how variable is "normal" gait, where in the
  cycle does that variation concentrate, and what are the dominant modes of
  variation?*
* **`dplyr::storms`** -- irregularly sampled position and intensity of every
  Atlantic tropical storm or hurricane from 1975-2022. *Question: do
  stronger storms travel further and move faster than weak ones?*

```{r packages, message = FALSE}
library(tf)
library(dplyr)
```

# 1. How variable is a "normal" gait cycle?

`gait` contains hip and knee angles, in degrees, for 39 boys, each evaluated
at 20 equally spaced points across one gait cycle. Wrap them into a single
two-component object:

```{r gait-mv}
data(gait)
g <- tfd_mv(list(hip = gait$hip_angle, knee = gait$knee_angle))
g
```

The two natural views of these 39 curves are the **time-series**
(`type = "facet"`, one panel per component) and the **trajectory in phase
space** (`type = "trajectory"`, the default for two-component objects).

```{r gait-facet, fig.height = 3.6}
plot(g, type = "facet", alpha = 0.4)
```

```{r gait-traj, fig.height = 4.2}
plot(g, alpha = 0.4)   # type = "trajectory" by default for d = 2
```

The facet view shows that both joints flex twice per cycle; the trajectory
view collapses time and reveals the characteristic "butterfly" loop traced
out as the leg cycles through stance and swing. Both views suggest the
between-subject spread is far from uniform along the cycle.

## Pointwise mean and standard deviation

`mean()` and `sd()` are vctrs group generics and dispatch component-wise:
they each return a length-1 `tfd_mv`.

```{r gait-mean-sd, fig.height = 3.8}
mu <- mean(g)
s  <- sd(g)

op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
for (k in seq_len(tf_ncomp(g))) {
  plot(tf_component(g, k), alpha = 0.25, ylab = names(tf_components(g))[k],
       main = names(tf_components(g))[k])
  arg <- tf_arg(mu)
  lines(arg, unlist(tf_evaluate(tf_component(mu, k), arg)), col = "red", lwd = 2)
  ev  <- unlist(tf_evaluate(tf_component(mu, k), arg))
  sdv <- unlist(tf_evaluate(tf_component(s,  k), arg))
  lines(arg, ev + 2 * sdv, col = "red", lwd = 1.2, lty = 2)
  lines(arg, ev - 2 * sdv, col = "red", lwd = 1.2, lty = 2)
}
par(op)
```

The pointwise sd peaks roughly where the angle itself is changing fastest --
around the stance/swing transitions -- so most of the between-subject
variability lives in *timing* of those transitions, not in the angles
attained at rest.

## Most- and least-varied subjects

`tf_arclength()` measures the path length traced out in `(hip, knee)`-space.
A short orbit = a subject whose `(hip, knee)` excursions are small or who
moves the two joints in lock-step; a long orbit = a subject with
large-amplitude, less synchronised excursions.

```{r gait-arc, fig.height = 4.2}
arc <- tf_arclength(g)
extreme <- c(which.min(arc), which.max(arc))
extreme

plot(g, alpha = 0.15)
lines(g[extreme[1]], col = "steelblue", lwd = 2.2)
lines(g[extreme[2]], col = "firebrick", lwd = 2.2)
legend("bottomright", bty = "n",
       legend = c(sprintf("min arc length (%.0f deg)", arc[extreme[1]]),
                  sprintf("max arc length (%.0f deg)", arc[extreme[2]])),
       col = c("steelblue", "firebrick"), lwd = 2)
```

## Modes of variation via FPC

`tfb_mv()` defaults to a spline basis; passing `basis = "fpc"` fits a
functional principal component basis per component. The reconstruction `g_b`
is itself a `tf_mv` and can be subtracted, plotted, or summarised like any
other:

```{r gait-fpc, fig.height = 3.8}
g_b <- tfb_mv(g, basis = "fpc", verbose = FALSE)

# how much of the per-component variance is captured by the leading few PCs?
fpc_var <- function(comp) attr(comp, "score_variance")
v_hip  <- fpc_var(tf_component(g_b, "hip"));  v_hip  <- v_hip  / sum(v_hip)
v_knee <- fpc_var(tf_component(g_b, "knee")); v_knee <- v_knee / sum(v_knee)

op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
barplot(head(v_hip,  6), names.arg = 1:6, main = "hip: variance share",
        ylab = "proportion", col = "grey70")
barplot(head(v_knee, 6), names.arg = 1:6, main = "knee: variance share",
        ylab = "proportion", col = "grey70")
par(op)

# residual = how well does a low-rank FPC basis approximate the raw curves?
g_round  <- vctrs::vec_cast(g_b, g)
resid    <- g - g_round
rmse_per_subject <- sqrt(unlist(lapply(tf_evaluations(resid),
                                       function(m) mean(m^2))))
summary(rmse_per_subject)
```

The first 2-3 FPCs explain almost all the variance in each joint, and the
RMSE between the raw and FPC-reconstructed `(hip, knee)` curves is on the
order of a degree -- consistent with a few interpretable modes (timing,
amplitude, asymmetry) accounting for almost all between-subject differences.

## Phase alignment with `tf_register`

Most of the pointwise variability above lives in *timing* (when does
heel-strike happen?), not amplitude. `tf_register()` on a `tf_mv` estimates
**one shared time-warp per curve** -- derived from a chosen reference
signal, here the hip angle -- and applies it to all components jointly.
That separates phase variability from amplitude variability:

```{r gait-register, fig.height = 4.4}
reg <- tf_register(g, method = "cc", ref_component = "hip")

g_aligned <- reg$registered
sd_b <- sd(g);         sd_a <- sd(g_aligned)
arg  <- tf_arg(sd_b)

op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
for (k in seq_len(tf_ncomp(g))) {
  nm <- names(tf_components(g))[k]
  yb <- unlist(tf_evaluate(tf_component(sd_b, k), arg))
  ya <- unlist(tf_evaluate(tf_component(sd_a, k), arg))
  plot(arg, yb, type = "l", lwd = 2, col = "grey40",
       ylim = c(0, max(yb, ya)),
       xlab = "cycle phase", ylab = paste("pointwise sd of", nm),
       main = nm)
  lines(arg, ya, lwd = 2, col = "firebrick")
  legend("topright", bty = "n", lwd = 2, col = c("grey40", "firebrick"),
         legend = c("raw", "phase-aligned"))
}
par(op)
```

Phase alignment knocks the largest sd peaks down by ~20-30 %: the
variability that remains is genuine between-subject *amplitude* difference,
the variability that was removed was just timing.

## Arc-length reparametrization

`tf_reparam_arclength()` traverses each curve at *unit speed in its value
space*. Visually the trajectory looks identical -- it's the same set of
points -- but the time axis has been redistributed so that equal time
intervals correspond to equal arc-length intervals. This is useful when
you care about the geometric *shape* of the orbit rather than how it was
traversed:

```{r gait-reparam, fig.height = 4.0}
g_unit <- tf_reparam_arclength(g)
sp_raw  <- tf_speed(g)
sp_unit <- tf_speed(g_unit)

# pointwise speed on the (FD-stable) interior of the cycle
tt <- seq(0.1, 0.9, length.out = 41)
mat_raw  <- vapply(seq_along(g), \(i) unlist(tf_evaluate(sp_raw[i],  tt)), numeric(41))
mat_unit <- vapply(seq_along(g), \(i) unlist(tf_evaluate(sp_unit[i], tt)), numeric(41))

op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
matplot(tt, mat_raw,  type = "l", lty = 1, col = grDevices::adjustcolor("steelblue", 0.4),
        xlab = "cycle phase", ylab = "speed (deg / cycle)",
        main = "raw parameterization")
matplot(tt, mat_unit, type = "l", lty = 1, col = grDevices::adjustcolor("firebrick", 0.4),
        xlab = "cycle phase", ylab = "speed (deg / cycle)",
        main = "arc-length parameterization")
par(op)
```

The raw plot shows large per-subject speed swings -- joints flex quickly
through the swing phase and stand almost still during stance. The
reparameterized plot is nearly flat: each curve is now traversed at the
same speed throughout the cycle. That speed is exactly the per-subject arc
length divided by the cycle duration -- the value we computed earlier.

# 2. Do stronger Atlantic storms travel further and faster?

`dplyr::storms` records the position (latitude / longitude) and intensity
(wind, pressure, category) of every Atlantic tropical storm or hurricane
from 1975-2022, sampled (mostly) every 6 hours along the storm's life.
Tracks have wildly different durations, so the data are genuinely
**irregularly sampled across storms** even though within-storm cadence is
regular -- a `tfd_irreg` per-storm structure on `(long, lat)`.

A naive `tfd_mv` on raw `(long, lat)` would let us do every analysis below,
but `tf_arclength` / `tf_speed` would report results in degrees / deg per
hour -- and longitude degrees aren't a physical unit: they shrink from
~111 km at the equator to ~78 km at 45 N. Atlantic hurricanes spend most of
their lives at moderate to high latitudes and *recurve to the northeast*,
so a deg-based metric would systematically overstate the apparent speed of
high-latitude motion. Project each storm into a local km coordinate system
first, using its own mean latitude as the projection reference:

```{r storms-build}
KM_PER_DEG <- 111.32

storms_clean <- storms |>
  mutate(
    storm_id = paste(name, year),
    ts       = as.POSIXct(ISOdate(year, month, day, hour), tz = "UTC")
  ) |>
  group_by(storm_id) |>
  # IBTrACS occasionally records a landfall row at the same time-stamp as a
  # regular 6-hourly obs; dedupe so `t_hours` is unique within a storm.
  distinct(ts, .keep_all = TRUE) |>
  mutate(
    t_hours = as.numeric(ts - min(ts), units = "hours"),
    ref_lat = mean(lat),                            # per-storm projection ref
    x_km    = (long - mean(long)) * KM_PER_DEG * cos(ref_lat * pi / 180),
    y_km    = (lat  - ref_lat)    * KM_PER_DEG
  ) |>
  filter(n() >= 16) |>                              # >= 4 days of life
  ungroup()

# peak Saffir-Simpson category per storm (0 = never reached hurricane status)
peak <- storms |>
  group_by(name, year) |>
  summarise(peak_cat = suppressWarnings(max(category, na.rm = TRUE)),
            .groups = "drop") |>
  mutate(peak_cat = ifelse(is.finite(peak_cat), as.integer(peak_cat), 0L),
         storm_id = paste(name, year))

# (long, lat) for the map view
tracks <- tfd_mv(list(
  long = tfd(storms_clean, id = "storm_id", arg = "t_hours", value = "long"),
  lat  = tfd(storms_clean, id = "storm_id", arg = "t_hours", value = "lat")
))

# (x_km, y_km) for physical-unit geometry
tracks_km <- tfd_mv(list(
  x = tfd(storms_clean, id = "storm_id", arg = "t_hours", value = "x_km"),
  y = tfd(storms_clean, id = "storm_id", arg = "t_hours", value = "y_km")
))

df <- tibble::tibble(
  storm_id = names(tracks),
  track    = tracks,
  track_km = tracks_km
) |>
  left_join(peak, by = "storm_id") |>
  mutate(strength = factor(
    pmin(peak_cat, 4),
    levels = 0:4,
    labels = c("TS/TD", "Cat 1", "Cat 2", "Cat 3", "Cat 4+")
  ))

df
```

## Track map by peak intensity

The geographic map uses the raw `(long, lat)` view (degrees on both axes,
as a reader would expect):

```{r storms-map, fig.height = 5.2}
pal <- c("grey70", "#fed976", "#feb24c", "#fd8d3c", "#e31a1c")
cols_per_storm <- pal[as.integer(df$strength)]

plot(df$track, alpha = 0.6, col = cols_per_storm,
     xlab = "longitude", ylab = "latitude")
legend("topleft", bty = "n", lwd = 2, col = pal, legend = levels(df$strength),
       title = "peak intensity")
```

Two visual populations: most tropical storms / weak hurricanes (grey /
yellow) stay in the Gulf of Mexico or run a short westward leg before
dissipating; major hurricanes (red) recurve into the mid-latitudes and
often run a long northeastward leg into the open ocean.

## Path length and forward speed in physical units

All geometric quantities below come from `tracks_km` (the projected, km
view), so they report in kilometres and km/h:

```{r storms-arclen, fig.height = 4}
df <- df |> mutate(
  path_km   = tf_arclength(track_km),
  duration  = vapply(tf_arg(track_km), \(t) max(t) - min(t), numeric(1)),
  mean_speed = path_km / duration              # km/h, lifetime average
)

boxplot(path_km ~ strength, data = df,
        xlab = "peak intensity", ylab = "track length (km)",
        col = pal, outline = FALSE, log = "y")
points(jitter(as.integer(df$strength), 0.4), df$path_km,
       pch = 16, col = grDevices::adjustcolor(cols_per_storm, alpha.f = 0.4))

df |> group_by(strength) |>
  summarise(n = dplyr::n(),
            median_path_km = round(median(path_km)),
            median_dur_hr  = median(duration),
            median_speed_kmh = round(median(mean_speed), 1))
```

Median path length grows sharply with peak intensity -- a Cat-4+ storm
typically traces 3-4x the path of a tropical storm -- driven both by
longer lifetimes *and* by higher mean forward speed.

## Forward speed time courses

The pointwise speed `||dr/dt||` (now in km/h) is a univariate `tfd`. Split
the speed envelope by peak intensity:

```{r storms-speed, fig.height = 4.2}
df <- df |> mutate(forward_speed = tf_speed(track_km))

op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
plot(df$forward_speed[df$strength == "TS/TD"], alpha = 0.35,
     ylim = c(0, 80), main = "TS / TD",
     xlab = "hours since genesis", ylab = "forward speed (km/h)")
plot(df$forward_speed[df$strength == "Cat 4+"], alpha = 0.6, col = "firebrick",
     ylim = c(0, 80), main = "Cat 4+",
     xlab = "hours since genesis", ylab = "forward speed (km/h)")
par(op)

df |> group_by(strength) |>
  summarise(median_lifetime_mean_speed_kmh = round(median(
              vapply(tf_evaluations(forward_speed),
                     \(v) mean(v, na.rm = TRUE), numeric(1))), 1),
            n = dplyr::n())
```

The Cat-4+ panel shows the canonical hurricane life-cycle in physical
units: slow westward drift at 10-20 km/h for the first few days, then a
sharp acceleration to 40-70 km/h during recurvature into the westerlies.
Weak storms rarely exceed 20-25 km/h.

## Smoothing the strongest tracks

`tfb_mv()` smooths each component into a spline basis. On the longest
6 tracks, the smoothed `(long, lat)` curves give a cleaner picture of the
overall track geometry without the 6-hourly sampling jitter:

```{r storms-tfb, fig.height = 4.8, warning = FALSE}
top6 <- df |> arrange(desc(path_km)) |> slice(1:6) |> pull(storm_id)
tb   <- tfb_mv(tracks[top6], k = 15, verbose = FALSE)

op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
plot(tracks[top6], col = 1:6, alpha = 0.7,
     xlab = "long", ylab = "lat", main = "raw observations")
plot(tb, col = 1:6,
     xlab = "long", ylab = "lat", main = "spline-smoothed")
par(op)
```

# 3. Recap

The two case studies above exercised exactly the same surface:

* construction from per-component `tf` vectors (`tfd_mv(list(...))`);
* `tf_components()`, `$component`, `tf_ncomp()` for component access;
* facet vs. trajectory plotting (`plot(..., type = ...)`);
* vctrs-native arithmetic / `mean()` / `sd()` returning a length-1 `tf_mv`;
* `tf_arclength()` and `tf_speed()` as geometric primitives on the bundle;
* `tfb_mv()` for smoothing (`basis = "spline"`) or PC decomposition
  (`basis = "fpc"`);
* `tf_register()` for joint warp estimation from a reference component, and
  `tf_reparam_arclength()` for unit-speed (shape-only) reparameterization;
* full dplyr / tibble integration via the underlying vctrs proxy.

Every numeric kernel here is the existing univariate one, mapped over the
`d` components by `vec_proxy.tf_mv`. Adding a new univariate `tf_*` verb
therefore extends to `tf_mv` automatically.

For the design rationale -- bundle-of-components vs. matrix-valued
evaluations vs. long/stacked vectors -- see `attic/design/multivariate.md`
in the package source.
