---
title: "Vector-valued functional data: gait cycles and hurricane tracks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vector-valued functional data: gait cycles and hurricane tracks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.2,
  dpi  = 96,
  out.width = "100%"
)
set.seed(1)
```

A `tf` vector represents a sample of functions `f: R -> R`. Many real-world
processes, though, produce several coupled signals that share a common argument
-- a `(hip, knee)` joint-angle pair sampled along a gait cycle, a `(longitude,
latitude)` storm track sampled in time, a `(x, y, z)` position recorded along
the body of a moving animal. The natural object is then a **vector-valued
function** `f: R -> R^d`, and `tf` represents a sample of those with the
`tf_mv` family:

* `tfd_mv` -- raw evaluations on a (regular or irregular) grid
* `tfb_mv` -- basis representations (spline or FPC)

Both are bundles of `d` ordinary `tf` components, so every numeric routine that
works on a univariate `tf` vector -- evaluation, arithmetic, summaries,
derivatives, integration, smoothing, basis fitting, registration -- carries
over component-wise without new code paths.

This article walks through two case studies:

1. **Gait data** (built into `tf`): regularly sampled hip / knee angles for 39
   boys. Shows construction, plotting, arithmetic, and basis fitting.
2. **Atlantic hurricane tracks** (from `dplyr::storms`): irregularly sampled
   longitude / latitude trajectories. Shows construction from a long
   data.frame, geometric primitives (arc length, speed), and dplyr integration.

```{r packages, message = FALSE}
library(tf)
library(dplyr)
```

# 1. Gait cycles

`gait` contains hip and knee angles, in degrees, for 39 boys, evaluated at 20
equally spaced points across one gait cycle. The two angle columns are already
ordinary `tfd` vectors:

```{r gait-data}
data(gait)
gait
```

A `tfd_mv` is built by passing a named list of `tf` vectors -- one per output
dimension -- to `tfd_mv()`:

```{r gait-mv}
g <- tfd_mv(list(hip = gait$hip_angle, knee = gait$knee_angle))
g
```

The print header summarises the structural pieces: `d = 2` components named
`(hip, knee)`, a common argument grid on `[0.025, 0.975]` (one gait cycle), and
per-component value ranges (`hip` in `[-12, 64]` degrees, `knee` in `[0, 82]`
degrees).

## Accessors

The component-wise design surfaces directly in the accessor API:

```{r gait-access}
tf_ncomp(g)
names(tf_components(g))
tf_domain(g)
g$hip                 # a plain tfd
identical(g$hip, gait$hip_angle)
```

`g[1:3]` subsets curves (length-preserving), `g$hip` and `tf_component(g, "knee")`
extract a single component as an ordinary univariate `tf`.

## Plotting

`plot()` on a `tf_mv` produces a small-multiples view, one panel per component,
sharing the argument axis:

```{r gait-plot, fig.height = 3.6}
plot(g)
```

For `d = 2` the geometric *trajectory* in the value space is often more
informative than the time-series view. `lines(g, type = "trajectory")` and
`plot(g, type = "trajectory")` switch to that:

```{r gait-traj, fig.height = 4.5}
plot(g, type = "trajectory", alpha = 0.5)
```

Each curve is one subject's `(hip(t), knee(t))` orbit over the gait cycle. The
loop reflects the coordinated swing-stance phasing.

## Arithmetic and summaries

All vctrs arithmetic, `Math`, and `Summary` group generics work
component-wise:

```{r gait-arith}
g_centered <- g - mean(g)
mean(g)
sd(g)
```

`mean(g)` is a length-1 `tfd_mv` -- the average gait cycle -- and `g - mean(g)`
is the centered sample.

## Fitting a smooth basis

`tfb_mv()` smooths each component independently into a basis expansion (spline
by default), returning a `tfb_mv`:

```{r gait-tfb}
gb <- tfb_mv(g, k = 15)
gb
```

Per-component basis options can be passed as component-named lists:

```{r gait-tfb-percomp, eval = FALSE}
tfb_mv(g, k = list(hip = 12, knee = 18), bs = list(hip = "cr", knee = "ps"))
```

Casting between `tfd_mv` and `tfb_mv` is symmetric and lossless up to the basis
quality:

```{r gait-roundtrip}
g_round <- vctrs::vec_cast(gb, g)  # tfb_mv -> tfd_mv on the original arg grid
max(abs(tf_evaluations(g - g_round)[[1]]))
```

# 2. Hurricane tracks

`dplyr::storms` records the position (latitude / longitude) and intensity
(wind, pressure, category) of every Atlantic tropical storm or hurricane from
1975 through 2022, sampled (mostly) every 6 hours along the storm's life.

```{r storms-peek}
storms
```

Two characteristics make this dataset interesting as a `tf_mv` example:

* Tracks have **different durations** -- from a couple of days to a few weeks
  -- so the data are genuinely *irregularly* sampled across storms even though
  the within-storm cadence is regular.
* Each row contains both spatial coordinates: `(long, lat)` is naturally a
  single `R^2`-valued function of time, not two separate functions.

Build a per-storm time-from-genesis variable, keep storms with at least 16
observations (4+ days of life), and assemble the `tfd_mv`:

```{r storms-mv}
tracks_long <- storms |>
  mutate(
    storm_id = paste(name, year),
    ts       = as.POSIXct(ISOdate(year, month, day, hour), tz = "UTC")
  ) |>
  group_by(storm_id) |>
  # IBTrACS occasionally records a landfall row at the same time-stamp as the
  # regular 6-hourly observation; drop those so `t_hours` is unique per storm.
  distinct(ts, .keep_all = TRUE) |>
  mutate(t_hours = as.numeric(ts - min(ts), units = "hours")) |>
  filter(n() >= 16) |>
  ungroup()

tracks <- tfd_mv(list(
  long = tfd(tracks_long, id = "storm_id", arg = "t_hours", value = "long"),
  lat  = tfd(tracks_long, id = "storm_id", arg = "t_hours", value = "lat")
))
tracks
```

The print header shows `long` and `lat` ranges spanning the North Atlantic
basin, and notes that the components have between 16 and ~100 evaluations each
-- the per-storm irregularity.

## Geographic trajectory plot

Plotting `(long(t), lat(t))` as a trajectory gives a map of the storm tracks:

```{r tracks-traj, fig.height = 4.5}
plot(tracks[1:50], type = "trajectory", alpha = 0.4,
     xlab = "longitude", ylab = "latitude")
```

The classic Atlantic basin shape emerges -- low-latitude westward tracks near
the Cape Verde Islands, recurvature into the mid-latitudes, and northeastward
tracks into the open ocean.

## Arc length: how far does a storm travel?

`tf_arclength()` computes the total path length traced out by each curve in
its value space. For lat/long coordinates that is (to first order) the great-
circle distance integrated along the track -- a proxy for how far the storm
travelled.

```{r tracks-arclength}
tracks_df <- tibble::tibble(
  storm_id = names(tracks),
  track    = tracks,
  path_deg = tf_arclength(tracks)
)
tracks_df |> arrange(desc(path_deg)) |> head(5)
```

Even though we built `tracks` from a list of irregular component vectors with
varying observed time spans, `tf_arclength()` handles the per-curve domain
clipping automatically.

## Speed and the segment view

`tf_speed(tracks)` returns the pointwise norm of the time derivative -- the
storm's forward speed in degrees/hour. The result is a length-`n` univariate
`tfd` and can be plotted, summarised, and joined back into the data frame
exactly like any other functional column:

```{r tracks-speed, fig.height = 4}
tracks_df <- tracks_df |>
  mutate(forward_speed = tf_speed(track))

plot(tracks_df$forward_speed[1:30], alpha = 0.4,
     xlab = "hours since genesis", ylab = "forward speed (deg/h)")
```

## Working inside tibbles with dplyr

`tf_mv` columns are full-fledged vctrs vectors, so they slot into tibbles and
flow through every dplyr verb:

```{r tracks-dplyr}
tracks_df |>
  mutate(
    decade   = 10 * (as.integer(sub(".* ", "", storm_id)) %/% 10),
    max_lat  = vapply(tf_evaluations(track), \(m) max(m[, "lat"]), numeric(1))
  ) |>
  group_by(decade) |>
  summarise(
    n_storms     = n(),
    mean_path    = mean(path_deg),
    mean_max_lat = mean(max_lat),
    .groups = "drop"
  )
```

`filter`, `arrange`, `slice`, `group_by` / `summarise`, `bind_rows`, and
`tidyr::nest` / `unnest` all keep the `tf_mv` column aligned with the rest of
the data frame.

## Fitting a smooth basis to a subset

The basis representation also works on this irregular sample. For brevity fit
a spline basis to a small subset of storms:

```{r tracks-tfb, warning = FALSE, fig.height = 4.5}
subset_ids <- tracks_df |>
  arrange(desc(path_deg)) |>
  slice(1:6) |>
  pull(storm_id)

tracks_b <- tfb_mv(tracks[subset_ids], k = 15, verbose = FALSE)
plot(tracks_b, type = "trajectory",
     xlab = "longitude", ylab = "latitude")
```

The smoothed trajectories highlight the gross shape of each track without the
6-hourly sampling jitter.

# Where to next

* The `tf_mv` family is *additive* on top of the univariate machinery: any new
  univariate `tf_*` verb you add automatically extends, component-wise, to
  `tf_mv` via the shared `vec_proxy` design.
* Tier-1 geometric primitives -- `tf_norm()`, `tf_speed()`, `tf_inner()`,
  `tf_distance()`, `tf_tangent()`, `tf_reparam_arclength()` -- are available
  for any `tf_mv` (and many work on univariate `tf` as well).
* Registration and warping (`tf_register`, `tf_warp`) treat `tf_mv` as a
  rigid bundle: a single warp is estimated from a chosen reference component
  (or a derived signal) and applied to all components in lockstep.

For the design rationale and a comparison with matrix-valued and long-format
alternatives, see `attic/design/multivariate.md` in the package source.
