---
title: "Vector-valued functional data: gait cycles and hurricane tracks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vector-valued functional data: gait cycles and hurricane tracks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.2,
  dpi  = 96,
  out.width = "100%"
)
set.seed(1)
```

A `tf` vector represents a sample of functions `f: R -> R`. Many real
measurement processes, though, produce several coupled signals that share a
common argument: the `(hip, knee)` joint-angle pair sampled across a gait
cycle, the `(longitude, latitude)` position of a hurricane sampled in time,
the `(x, y, z)` body coordinates of a moving animal. The natural object is a
**vector-valued function** `f: R -> R^d`, and `tf` represents a sample of
those with the `tf_mv` family: `tfd_mv` for raw evaluations and `tfb_mv` for
basis representations. Internally a `tf_mv` is just a bundle of `d` ordinary
`tf` vectors, so every univariate verb -- evaluation, arithmetic, smoothing,
basis fitting, derivatives, integration -- extends component-wise for free.

This article uses two real datasets to put the API through its paces:

* **`tf::gait`** -- regularly sampled hip and knee angles for 39 boys across
  one gait cycle. *Question: how variable is "normal" gait, where in the
  cycle does that variation concentrate, and what are the dominant modes of
  variation?*
* **`dplyr::storms`** -- irregularly sampled position and intensity of every
  Atlantic tropical storm or hurricane from 1975-2022. *Question: do
  stronger storms travel further and move faster than weak ones?*

```{r packages, message = FALSE}
library(tf)
library(dplyr)
```

# 1. How variable is a "normal" gait cycle?

`gait` contains hip and knee angles, in degrees, for 39 boys, each evaluated
at 20 equally spaced points across one gait cycle. Wrap them into a single
two-component object:

```{r gait-mv}
data(gait)
g <- tfd_mv(list(hip = gait$hip_angle, knee = gait$knee_angle))
g
```

The two natural views of these 39 curves are the **time-series**
(`type = "facet"`, one panel per component) and the **trajectory in phase
space** (`type = "trajectory"`, the default for two-component objects).

```{r gait-facet, fig.height = 3.6}
plot(g, type = "facet", alpha = 0.4)
```

```{r gait-traj, fig.height = 4.2}
plot(g, alpha = 0.4)   # type = "trajectory" by default for d = 2
```

The facet view shows that both joints flex twice per cycle; the trajectory
view collapses time and reveals the characteristic "butterfly" loop traced
out as the leg cycles through stance and swing. Both views suggest the
between-subject spread is far from uniform along the cycle.

## Pointwise mean and standard deviation

`mean()` and `sd()` are vctrs group generics and dispatch component-wise:
they each return a length-1 `tfd_mv`. Plot the raw curves with `plot.tf`,
then overlay the mean +/- 2 sd envelope with `lines.tf` -- all arithmetic
on the components is component-wise on `tf_mv`, so `mu + 2 * s` is itself
a `tfd_mv` and `tf_component(...)` gives the per-axis envelope:

```{r gait-mean-sd, fig.height = 3.8}
mu <- mean(g)
s  <- sd(g)

op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
for (k in seq_len(tf_ncomp(g))) {
  nm <- names(tf_components(g))[k]
  plot(tf_component(g, k), alpha = 0.25,
       ylab = nm, main = nm)
  lines(tf_component(mu, k),              col = "red", lwd = 2)
  lines(tf_component(mu + 2 * s, k),      col = "red", lwd = 1.2, lty = 2)
  lines(tf_component(mu - 2 * s, k),      col = "red", lwd = 1.2, lty = 2)
}
par(op)
```

The pointwise sd peaks roughly where the angle itself is changing fastest --
around the stance/swing transitions -- so most of the between-subject
variability lives in *timing* of those transitions, not in the angles
attained at rest.

## Most- and least-varied subjects

`tf_arclength()` measures the path length traced out in `(hip, knee)`-space.
A short orbit = a subject whose `(hip, knee)` excursions are small or who
moves the two joints in lock-step; a long orbit = a subject with
large-amplitude, less synchronised excursions.

```{r gait-arc, fig.height = 4.2}
arc <- tf_arclength(g)
extreme <- c(which.min(arc), which.max(arc))
extreme

plot(g, alpha = 0.15)
lines(g[extreme[1]], col = "steelblue", lwd = 2.2)
lines(g[extreme[2]], col = "firebrick", lwd = 2.2)
legend("bottomright", bty = "n",
       legend = c(sprintf("min arc length (%.0f deg)", arc[extreme[1]]),
                  sprintf("max arc length (%.0f deg)", arc[extreme[2]])),
       col = c("steelblue", "firebrick"), lwd = 2)
```

## Phase alignment with `tf_register`

The pointwise sd above conflates two very different kinds of
between-subject variability: *amplitude* (how high does the knee swing?)
and *phase* (at what fraction of the cycle does heel-strike happen?).
`tf_register()` on a `tf_mv` estimates **one shared time-warp per curve**
and applies it to all components jointly. The warp is derived from a
single univariate "reference signal" -- which component (or which derived
quantity) to use is the central modelling choice.

Three reasonable choices, each emphasising different events in the cycle:

```{r gait-register-fit, results = "hide"}
r_hip   <- tf_register(g, method = "cc", ref_component = "hip")
r_knee  <- tf_register(g, method = "cc", ref_component = "knee")
# any univariate function of x can serve as the reference signal -- here the
# tangential *speed* in (hip, knee)-space, combining both components
r_speed <- tf_register(g, method = "cc", ref_component = tf_speed)
```

### Aligned trajectories

`plot(..., type = "facet")` shows the aligned curves for each choice of
reference signal. The reference signal that was *used* to register a
sample shrinks most strongly, but the partner component can shrink or
expand depending on how phase-coupled the two are:

```{r gait-register-traj, fig.width = 8.2, fig.height = 7.2}
plot_aligned <- function(reg, title) {
  plot(reg$registered, type = "facet", alpha = 0.3)
  mtext(title, side = 3, line = -1.3, outer = TRUE, font = 2)
}

op <- par(mfrow = c(3, 1), oma = c(0, 0, 0, 0))
plot_aligned(r_hip,   "ref = hip angle")
plot_aligned(r_knee,  "ref = knee angle")
plot_aligned(r_speed, "ref = speed = ||(hip', knee')||")
par(op)
```

The "ref = knee" panel visibly tightens both components -- the knee carries
the sharper heel-strike signal that the alignment can lock onto, and the
hip (which is phase-coupled to it) tightens for free. The hip-only and
speed-driven alignments knock down their target signal but leave the
partner component largely untouched.

### Estimated warping functions

The warps themselves are univariate `tfd`s (one per subject), mapping the
raw cycle phase to the aligned phase. `plot.tf` handles them directly:

```{r gait-register-warps, fig.width = 8.2, fig.height = 3.6}
warps_hip   <- tf_invert(r_hip$inv_warps)
warps_knee  <- tf_invert(r_knee$inv_warps)
warps_speed <- tf_invert(r_speed$inv_warps)

op <- par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
plot(warps_hip,   alpha = 0.5, main = "warp from hip",   ylab = "aligned phase")
abline(0, 1, lty = 3)
plot(warps_knee,  alpha = 0.5, main = "warp from knee",  ylab = "aligned phase")
abline(0, 1, lty = 3)
plot(warps_speed, alpha = 0.5, main = "warp from speed", ylab = "aligned phase")
abline(0, 1, lty = 3)
par(op)
```

The dotted diagonal is the identity warp (no time distortion). The
spread around the diagonal tells you the magnitude and pattern of the
phase variability captured by each reference signal: warps from `knee`
and `speed` bunch tightly near identity in the first half of the cycle
and fan out around the swing phase, where heel-strike timing varies
between subjects. The `hip`-driven warps are noticeably broader
elsewhere -- the hip signal is too smooth to lock the alignment to a
single sharp event.

### Quantifying the alignment

Compare the maximum pointwise sd before and after each registration:

```{r gait-register-table}
peak_sd <- function(f, k) {
  arg <- tf_arg(f)
  max(unlist(tf_evaluate(tf_component(sd(f), k), arg)))
}
out <- data.frame(
  reference = c("raw (no reg)", "hip", "knee", "speed"),
  sd_hip = c(peak_sd(g, "hip"),
             peak_sd(r_hip$registered,  "hip"),
             peak_sd(r_knee$registered, "hip"),
             peak_sd(r_speed$registered,"hip")),
  sd_knee = c(peak_sd(g, "knee"),
              peak_sd(r_hip$registered,  "knee"),
              peak_sd(r_knee$registered, "knee"),
              peak_sd(r_speed$registered,"knee"))
)
out
```

`knee` is the most informative reference signal here, both because the
knee has the cycle's strongest event (heel-strike) and because that event
phase-couples to hip flexion. We carry the knee-aligned sample forward.

## Modes of variation via FPC

With phase variability removed, fit a per-component FPC basis. The
leading PCs now describe *amplitude* modes only:

```{r gait-fpc, fig.height = 3.8}
g_aligned <- r_knee$registered
g_b <- tfb_mv(g_aligned, basis = "fpc", verbose = FALSE)

fpc_var <- function(comp) attr(comp, "score_variance")
v_hip  <- fpc_var(tf_component(g_b, "hip"));  v_hip  <- v_hip  / sum(v_hip)
v_knee <- fpc_var(tf_component(g_b, "knee")); v_knee <- v_knee / sum(v_knee)

op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
barplot(head(v_hip,  6), names.arg = 1:6, main = "hip: variance share",
        ylab = "proportion", col = "grey70")
barplot(head(v_knee, 6), names.arg = 1:6, main = "knee: variance share",
        ylab = "proportion", col = "grey70")
par(op)

# residual: how well does the low-rank FPC basis approximate the curves?
g_round <- vctrs::vec_cast(g_b, g_aligned)
resid   <- g_aligned - g_round
rmse_per_subject <- sqrt(unlist(lapply(tf_evaluations(resid),
                                       function(m) mean(m^2))))
summary(rmse_per_subject)
```

The first 2-3 FPCs explain almost all the amplitude variance in each
joint, and the RMSE between the registered and FPC-reconstructed
`(hip, knee)` curves is on the order of a degree.

## Arc-length reparametrization

`tf_reparam_arclength()` traverses each curve at *unit speed in its value
space*. The trajectory looks identical -- it's the same set of points --
but the time axis has been redistributed so that equal time intervals
correspond to equal arc-length intervals. `tf_speed()` returns a plain
`tfd`, so `plot.tf` and `lines.tf` work directly:

```{r gait-reparam, fig.height = 4.0}
g_unit  <- tf_reparam_arclength(g)
sp_raw  <- tf_speed(g)
sp_unit <- tf_speed(g_unit)

op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
plot(sp_raw,  alpha = 0.4, ylim = c(0, max(c(
       unlist(tf_evaluations(sp_raw)),
       unlist(tf_evaluations(sp_unit))), na.rm = TRUE)),
     main = "raw parameterization", ylab = "speed (deg / cycle)")
lines(mean(sp_raw),  col = "firebrick", lwd = 2)
plot(sp_unit, alpha = 0.4, ylim = c(0, max(c(
       unlist(tf_evaluations(sp_raw)),
       unlist(tf_evaluations(sp_unit))), na.rm = TRUE)),
     main = "arc-length parameterization", ylab = "speed (deg / cycle)")
lines(mean(sp_unit), col = "firebrick", lwd = 2)
par(op)
```

The raw plot shows large per-subject speed swings -- joints flex quickly
through swing and stand almost still during stance. The reparameterized
plot is approximately flat: each curve is traversed at the same speed
throughout the cycle, equal to its per-subject arc length divided by the
cycle duration.

# 2. Atlantic storms as 4-dimensional curves

`dplyr::storms` records four time-varying quantities for every Atlantic
tropical storm or hurricane from 1975-2022, sampled every 6 hours along
the storm's life: position (longitude, latitude) and intensity (sustained
wind speed, central pressure). The natural object is a single
**four-component** vector-valued curve per storm, `f_i: [0, T_i] -> R^4`
-- spatial trajectory and intensity life-cycle bundled together.

Two practical points:

* `tracks_km` -- the *spatial* `(x_km, y_km)` view in physical units
  (projecting each storm's `long` / `lat` into a per-storm local-km
  frame at its own mean latitude) -- is what we use for arc length and
  forward speed, because deg/h speeds wrongly upweight east-west motion
  at high latitudes (a longitude degree is ~111 km at the equator but
  only ~78 km at 45 N).
* Most life-cycle comparisons between short and long storms only make
  sense on a **normalised time axis** `phase = t / T_i in [0, 1]`. We
  build a normalised 4-d object on top of the real-time one and use the
  appropriate version per question.

```{r storms-build}
KM_PER_DEG <- 111.32

storms_clean <- storms |>
  mutate(
    storm_id = paste(name, year),
    ts       = as.POSIXct(ISOdate(year, month, day, hour), tz = "UTC")
  ) |>
  group_by(storm_id) |>
  distinct(ts, .keep_all = TRUE) |>      # dedupe duplicate-timestamp rows
  mutate(
    t_hours = as.numeric(ts - min(ts), units = "hours"),
    phase   = if (max(t_hours) > 0) t_hours / max(t_hours) else 0,
    ref_lat = mean(lat),
    x_km    = (long - mean(long)) * KM_PER_DEG * cos(ref_lat * pi / 180),
    y_km    = (lat  - ref_lat)    * KM_PER_DEG
  ) |>
  filter(n() >= 16, !is.na(wind), !is.na(pressure)) |>  # >= 4 days
  ungroup()

# spatial (km, real time)
tracks_km <- tfd_mv(list(
  x = tfd(storms_clean, id = "storm_id", arg = "t_hours", value = "x_km"),
  y = tfd(storms_clean, id = "storm_id", arg = "t_hours", value = "y_km")
))

# full 4-d, on normalised life-cycle phase
tracks4 <- tfd_mv(list(
  long = tfd(storms_clean, id = "storm_id", arg = "phase", value = "long"),
  lat  = tfd(storms_clean, id = "storm_id", arg = "phase", value = "lat"),
  wind = tfd(storms_clean, id = "storm_id", arg = "phase", value = "wind"),
  pres = tfd(storms_clean, id = "storm_id", arg = "phase", value = "pressure")
))
tracks4

# storm-level metadata
peak <- storms |>
  group_by(name, year) |>
  summarise(peak_cat = suppressWarnings(max(category, na.rm = TRUE)),
            .groups = "drop") |>
  mutate(peak_cat = ifelse(is.finite(peak_cat), as.integer(peak_cat), 0L),
         storm_id = paste(name, year))

df <- tibble::tibble(
  storm_id  = names(tracks4),
  track     = tracks4,
  track_km  = tracks_km
) |>
  left_join(peak, by = "storm_id") |>
  mutate(strength = factor(
    pmin(peak_cat, 4),
    levels = 0:4,
    labels = c("TS/TD", "Cat 1", "Cat 2", "Cat 3", "Cat 4+")
  ))
```

## A single 4-d curve: Hurricane Katrina (2005)

For a single storm the 4-component `tf_mv` is naturally displayed
in facet mode (`type = "facet"`) -- one panel per component, all on the
same normalised cycle phase axis:

```{r storms-katrina, fig.width = 8, fig.height = 5.4}
katrina <- df |> filter(storm_id == "Katrina 2005") |> pull(track)
plot(katrina, type = "facet", lwd = 2)
```

Wind and pressure are anti-correlated -- pressure dips below 920 mbar as
wind peaks above 150 knots near phase ~0.55 -- and the `(long, lat)`
panels show Katrina sliding northwest across the Gulf and turning north
into Louisiana. Bundling intensity and trajectory as a single `tfd_mv`
lets every operation downstream (subsetting, summaries, arithmetic,
plotting) treat them as one object.

## Track map, faceted by peak intensity

The geographic view uses the `(long, lat)` components only. Faceting by
peak Saffir-Simpson category separates the populations cleanly, and a
coastline backdrop (from `maps::map`) anchors the geography:

```{r storms-map, fig.width = 8.2, fig.height = 6.8}
have_maps <- requireNamespace("maps", quietly = TRUE)
draw_coast <- function(xlim, ylim) {
  if (!have_maps) return(invisible())
  m <- maps::map("world", plot = FALSE,
                 xlim = xlim, ylim = ylim, fill = FALSE)
  lines(m$x, m$y, col = "grey55", lwd = 0.6)
}

xlim <- range(unlist(tf_evaluations(tf_component(df$track, "long"))), na.rm = TRUE)
ylim <- range(unlist(tf_evaluations(tf_component(df$track, "lat"))),  na.rm = TRUE)

op <- par(mfrow = c(2, 3), mar = c(3.4, 3.4, 2, 1), mgp = c(2, 0.7, 0))
pal <- c("grey50", "#fed976", "#feb24c", "#fd8d3c", "#e31a1c")
for (k in seq_along(levels(df$strength))) {
  lev   <- levels(df$strength)[k]
  trks  <- df$track[df$strength == lev]
  spatial <- tfd_mv(list(
    long = tf_component(trks, "long"),
    lat  = tf_component(trks, "lat")
  ))
  plot(range(xlim), range(ylim), type = "n",
       xlab = "long", ylab = "lat",
       main = sprintf("%s (n = %d)", lev, length(trks)))
  draw_coast(xlim, ylim)
  lines(spatial, col = pal[k], alpha = 0.4, lwd = 1)
}
plot.new()
legend("center", bty = "n", lwd = 2, col = pal, legend = levels(df$strength),
       title = "peak intensity", cex = 1.05)
par(op)
```

Tropical storms / depressions stay in the southern, western basin; cat-3
and cat-4+ tracks reach further north and east -- they live longer, get
caught by the westerlies, and recurve out to sea.

## Scalar features extracted from the 4-d object

Every component of a `tf_mv` is a real `tf` vector and supports the full
univariate API, so per-storm scalar features fall out as one-liners:

```{r storms-scalars}
df <- df |> mutate(
  path_km    = tf_arclength(track_km),
  duration   = vapply(tf_arg(track_km), \(t) max(t) - min(t), numeric(1)),
  mean_speed = path_km / duration,            # km/h, lifetime average
  peak_wind  = vapply(tf_evaluations(tf_component(track, "wind")), max, numeric(1)),
  min_pres   = vapply(tf_evaluations(tf_component(track, "pres")), min, numeric(1))
)

df |>
  group_by(strength) |>
  summarise(
    n = dplyr::n(),
    median_path_km     = round(median(path_km)),
    median_speed_kmh   = round(median(mean_speed), 1),
    median_peak_wind   = median(peak_wind),
    median_min_pres    = median(min_pres),
    .groups = "drop"
  )
```

Median path length grows roughly 4x from tropical-storm to cat-4+ storms,
median forward speed roughly doubles, and the canonical wind / pressure
gap between categories is visible.

## Intensity and forward-speed life-cycles per category

Plotting all forward-speed and intensity time courses at once is hopeless
(hundreds of irregular noodles). What we actually want is the
*mean curve per intensity stratum* on the normalised lifecycle phase
axis. Evaluate each component on a common phase grid, average within each
stratum, and re-wrap the result as a length-`G` univariate `tfd` so we
can use the standard `plot.tf` / `lines.tf` machinery:

```{r storms-lifecycle, fig.width = 8.2, fig.height = 6.8}
phase_grid <- seq(0, 1, length.out = 41)

# forward speed on normalised time: re-arg each storm's km-speed by
# t / T_i, so all storms share phase domain [0, 1].
speed_km  <- tf_speed(df$track_km)
durations <- df$duration
speed_long <- do.call(rbind, lapply(seq_along(speed_km), function(i) {
  data.frame(
    id    = names(speed_km)[i],
    phase = tf_arg(speed_km[i])[[1]] / durations[i],
    value = tf_evaluations(speed_km[i])[[1]]
  )
}))
speed_phase <- tfd(speed_long, id = "id", arg = "phase", value = "value",
                   domain = c(0, 1))

# per-stratum mean curve on the regular phase grid, packaged as length-G tfd
stratum_mean_tfd <- function(comp, grp, grid = phase_grid) {
  mat <- as.matrix(comp, arg = grid, interpolate = TRUE)
  means <- vapply(levels(grp), function(g) {
    rows <- which(grp == g)
    if (length(rows) < 2L) return(rep(NA_real_, length(grid)))
    colMeans(mat[rows, , drop = FALSE], na.rm = TRUE)
  }, numeric(length(grid)))
  out <- tfd(t(means), arg = grid, domain = c(0, 1))
  names(out) <- levels(grp)
  out
}

wind_avg  <- stratum_mean_tfd(tf_component(df$track, "wind"), df$strength)
pres_avg  <- stratum_mean_tfd(tf_component(df$track, "pres"), df$strength)
speed_avg <- stratum_mean_tfd(speed_phase,                     df$strength)

op <- par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
plot(wind_avg, col = pal, lwd = 2, alpha = 1,
     xlab = "lifecycle phase", ylab = "wind (knots)",
     main = "mean sustained wind")
legend("topright", bty = "n", lwd = 2, col = pal, legend = levels(df$strength),
       cex = 0.9, title = "peak intensity")
plot(pres_avg,  col = pal, lwd = 2, alpha = 1,
     xlab = "lifecycle phase", ylab = "pressure (mbar)",
     main = "mean central pressure")
plot(speed_avg, col = pal, lwd = 2, alpha = 1,
     xlab = "lifecycle phase", ylab = "forward speed (km/h)",
     main = "mean forward speed")
plot.new()
text(0.5, 0.7, "wind and pressure peak\nnear lifecycle phase 0.5;",
     adj = 0.5, cex = 1.05)
text(0.5, 0.35, "forward speed peaks\nlate, during recurvature",
     adj = 0.5, cex = 1.05)
par(op)
```

Three clean stories on a single `tf_mv` object: intensity peaks
mid-life-cycle (around `phase ~ 0.5`) and the gap between categories is
roughly uniform across the cycle; central pressure mirrors wind; forward
speed peaks much later (around `phase ~ 0.75`) and the strongest storms
accelerate the most -- the canonical recurvature signature.

## Smoothing on normalised time

Fitting a basis representation on raw real-time data has a tractable
domain-extrapolation problem -- the longest storm runs 600+ hours, but
most tracks end after 100-200 hours, and a shared basis over [0, 600]
extrapolates wildly. On the **normalised phase** version this just
disappears, because every storm spans `[0, 1]`:

```{r storms-tfb, fig.width = 8.2, fig.height = 4.8, warning = FALSE}
top6 <- df |> arrange(desc(path_km)) |> slice(1:6) |> pull(storm_id)

# fit a per-component spline basis on (long, lat, wind, pres) over [0, 1]
tb <- tfb_mv(df$track[df$storm_id %in% top6], k = 12, verbose = FALSE)

# pull just the (long, lat) components out of the 4-d objects so plot.tf_mv
# defaults to the trajectory (long, lat) view
raw_xy <- tfd_mv(tf_components(df$track[df$storm_id %in% top6])[c("long","lat")])
sm_xy  <- tfd_mv(tf_components(tb)[c("long","lat")])

op <- par(mfrow = c(1, 2), mar = c(3.6, 3.6, 2, 1), mgp = c(2, 0.7, 0))
plot(range(xlim), range(ylim), type = "n",
     xlab = "long", ylab = "lat", main = "raw observations")
draw_coast(xlim, ylim)
lines(raw_xy, col = 1:6, lwd = 1.6)
plot(range(xlim), range(ylim), type = "n",
     xlab = "long", ylab = "lat", main = "spline-smoothed")
draw_coast(xlim, ylim)
lines(sm_xy, col = 1:6, lwd = 1.6)
par(op)
```

The smoothed trajectories pick up the gross recurvature shape without the
6-hourly sampling jitter and -- thanks to the normalised time axis --
without any extrapolation pathology. The same `tb` object also carries
clean smoothed wind / pressure curves for each storm; for example:

```{r storms-tfb-intensity, fig.width = 8.2, fig.height = 3.4}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
plot(tf_component(tb, "wind"), col = 1:6, lwd = 2,
     xlab = "lifecycle phase", ylab = "wind (knots)",
     main = "smoothed wind")
plot(tf_component(tb, "pres"), col = 1:6, lwd = 2,
     xlab = "lifecycle phase", ylab = "pressure (mbar)",
     main = "smoothed pressure")
par(op)
```

# 3. Recap

The two case studies above exercised exactly the same surface:

* construction from per-component `tf` vectors (`tfd_mv(list(...))`);
* `tf_components()`, `$component`, `tf_ncomp()` for component access;
* facet vs. trajectory plotting (`plot(..., type = ...)`);
* vctrs-native arithmetic / `mean()` / `sd()` returning a length-1 `tf_mv`;
* `tf_arclength()` and `tf_speed()` as geometric primitives on the bundle;
* `tfb_mv()` for smoothing (`basis = "spline"`) or PC decomposition
  (`basis = "fpc"`);
* `tf_register()` for joint warp estimation from a reference component, and
  `tf_reparam_arclength()` for unit-speed (shape-only) reparameterization;
* full dplyr / tibble integration via the underlying vctrs proxy.

Every numeric kernel here is the existing univariate one, mapped over the
`d` components by `vec_proxy.tf_mv`. Adding a new univariate `tf_*` verb
therefore extends to `tf_mv` automatically.

For the design rationale -- bundle-of-components vs. matrix-valued
evaluations vs. long/stacked vectors -- see `attic/design/multivariate.md`
in the package source.
